---
title: "Data downloads"
subtitle: "Every snapshot the trackers consume, available as parquet"
toc: false
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(here)
  library(jsonlite)
  library(htmltools)
})
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))

apis_dir <- here::here("data-apis")
final_dir <- here::here("data-final")

# Repo URL prefix that GitHub uses for raw-blob downloads. Each snapshot
# in data-apis/ is committed nightly, so the raw blob URL always points
# at the most-recent successful run.
repo_raw <- "https://github.com/michalovadek/euverse/raw/main"
repo_blob <- "https://github.com/michalovadek/euverse/blob/main"

# Build a table row per snapshot in data-apis/ + data-final/.
snapshots_in <- function(dir_label, dir_path, raw_prefix, blob_prefix) {
  files <- list.files(dir_path, pattern = "\\.parquet$", full.names = TRUE)
  if (length(files) == 0L) return(NULL)
  tibble::tibble(
    folder    = dir_label,
    file      = basename(files),
    bytes     = file.info(files)$size,
    raw_url   = paste0(raw_prefix, "/", dir_label, "/", basename(files)),
    blob_url  = paste0(blob_prefix, "/", dir_label, "/", basename(files))
  )
}

# Read the meta sidecar if it exists; otherwise return placeholders.
attach_meta <- function(row) {
  meta_path <- file.path(here::here(row$folder),
                         sub("\\.parquet$", ".meta.json", row$file))
  if (!file.exists(meta_path)) {
    row$source <- "(no meta)"; row$fetched_at <- NA; row$rows <- NA
    return(row)
  }
  meta <- jsonlite::fromJSON(meta_path)
  row$source     <- as.character(meta$source %||% meta$src %||% "(unknown)")
  row$fetched_at <- as.character(meta$fetched_at %||% meta$built_at %||% NA)
  row$rows       <- as.integer(meta$rows %||% NA)
  row
}
`%||%` <- function(x, y) if (is.null(x) || (length(x) == 1 && is.na(x))) y else x

snapshots <- bind_rows(
  snapshots_in("data-apis",  apis_dir,  repo_raw, repo_blob),
  snapshots_in("data-final", final_dir, repo_raw, repo_blob)
) |>
  group_split(row_number()) |>
  lapply(attach_meta) |>
  bind_rows() |>
  arrange(folder, source)
```

Every chart on this site reads from a parquet snapshot in
[`data-apis/`](`r repo_blob`/data-apis) or
[`data-final/`](`r repo_blob`/data-final). The snapshots are committed
to the repository on every nightly run, so the raw-blob URLs in the
table below always point to the most-recent successful render.

Use these if you want to do your own analysis on the same data the
trackers are showing, without having to re-query Eur-Lex / Eurostat /
the ECB yourself.

```{r snapshots-table, echo=FALSE}
snapshots |>
  transmute(
    Folder    = folder,
    Source    = source,
    `Last fetched (UTC)` = ifelse(is.na(fetched_at), "—",
                                  format(as.POSIXct(fetched_at, tz = "UTC",
                                                    format = "%Y-%m-%dT%H:%M:%SZ"),
                                         "%Y-%m-%d %H:%M")),
    Rows      = ifelse(is.na(rows), "?", format(rows, big.mark = ",")),
    Size      = ifelse(bytes < 1024^2,
                       paste0(round(bytes / 1024, 1), " KB"),
                       paste0(round(bytes / 1024^2, 1), " MB")),
    Download  = sprintf('<a href="%s" target="_blank">parquet</a> / <a href="%s" target="_blank">view</a>',
                        raw_url, blob_url)
  ) |>
  dt_mo(escape = FALSE)
```

# How to read these locally

```r
install.packages("arrow")                    # one-time
df <- arrow::read_parquet(
  "https://github.com/michalovadek/euverse/raw/main/data-apis/eurlex_acts_latest.parquet"
)
```

Or with Python + `pyarrow`:

```python
import pyarrow.parquet as pq
import urllib.request
url = "https://github.com/michalovadek/euverse/raw/main/data-apis/eurlex_acts_latest.parquet"
urllib.request.urlretrieve(url, "/tmp/acts.parquet")
df = pq.read_table("/tmp/acts.parquet").to_pandas()
```

# Schema and freshness contract

Each `<src>_latest.parquet` ships with a `<src>_latest.meta.json`
sidecar carrying `{source, fetched_at, rows, schema_hash}`. The
`schema_hash` is a SHA-256 of the column names; if it changes between
two consecutive snapshots, the source has reshaped its schema and any
downstream code reading the parquet should be reviewed.

If a nightly fetcher fails, **the snapshot stays at the previous run's
state** — pages keep rendering, the freshness badge turns red, and
the next morning's fetch tries again. There is no historical archive:
re-query the upstream source if you need older data.
