---
title: "euverse"
subtitle: "Public-data dashboards on European Union institutions, refreshed every night"
toc: false
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here)
  library(htmltools)
})

# Build one tracker card as a single un-indented HTML string. We use raw
# sprintf rather than htmltools::div() because htmltools indents nested
# children by 2 spaces and Pandoc's markdown reader treats indented HTML
# blocks as code, wrapping the rendered card content in <pre><code> on
# the live site. Single-line output dodges that entirely.
card <- function(href, title, blurb) {
  esc <- htmltools::htmlEscape
  paste0(
    '<div class="tracker-card">',
    '<a href="', esc(href), '" class="stretched-link"><h3>', esc(title), '</h3></a>',
    '<p class="tracker-blurb">', esc(blurb), '</p>',
    '</div>'
  )
}
```

The European Union generates an extraordinary amount of public data:
court rulings, vetoes, legal acts, financial indicators. Most of it lives
behind unfriendly APIs and PDF archives. This site pulls those streams
into plain-language dashboards that re-render every night.

::: {.tracker-grid}

```{r cards, echo=FALSE, results='asis'}
cat(
  card("trackers/eu-finance.html", "EU Finance",
       "Bond yields, inflation and the ECB's key rate, plotted against the post-2008 sovereign-debt back-history."),
  card("trackers/eu-vetoes.html", "EU Vetoes",
       "Every Member State veto on a Council file since 2010, hand-curated from the public record."),
  card("trackers/eu-court.html", "EU Court",
       "Caseload, procedure types, rapporteur productivity and country-of-origin maps for the CJEU."),
  card("trackers/eu-law.html", "EU Law",
       "Regulations, directives, decisions and recommendations adopted by the EU since 1952, plus how long they took to pass."),
  card("trackers/eu-enlargement.html", "EU Enlargement",
       "Membership of the EU from the Founding Six in 1957 to today, traced wave by wave (and the one withdrawal)."),
  sep = "\n"
)
```

:::

## How it works

Each tracker rebuilds itself overnight: a scheduled job re-pulls the
underlying data and the page re-renders, so the charts reflect what the
sources say that day. Most of the data comes from public APIs,
complemented on occasion by additional external or hand-curated sources.
When a source is unreachable the page falls back to the last good copy
and shows the date the data was fetched, so you can see how current the
numbers are.

The full source is on
[GitHub](https://github.com/michalovadek/euverse).

## About

I'm [Michal Ovádek](https://michalovadek.github.io/), a Lecturer in
European Institutions, Politics and Policy at UCL's Department of
Political Science. My research focuses on European integration, judicial
politics and Central and Eastern Europe. Before UCL I held research
positions in Sweden and Belgium and worked as a political advisor in the
European Parliament.

Cite individual trackers by their printed citation block; cite the project
as a whole as: Ovádek, M. *euverse*. Accessed `r format(Sys.Date(), "%d %B %Y")`.
Available at <https://michalovadek.github.io/euverse/>.
