---
title: "euverse"
subtitle: "Public-data dashboards on European Union institutions, refreshed every night"
toc: false
page-layout: full
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(here)
  library(htmltools)
})
source(here::here("R", "freshness.R"))

# Load a few minimal facts from each tracker's snapshot so the landing
# page surfaces what the trackers actually contain instead of generic
# "EU stats here" copy. Each fact is the single line a reader would
# want to know before clicking through.

# ---- EU Finance: latest EA bond yield ---------------------------------------
yields    <- read_with_freshness("eurostat_teimf050")
ea_latest <- yields$data |>
  filter(geo == "EA") |>
  arrange(desc(date)) |>
  slice(1)
fin_fact <- sprintf(
  "Euro Area 10-year bond yield at %s%%",
  format(round(ea_latest$yield_pct, 2), nsmall = 2)
)
fin_date <- format(as.Date(ea_latest$date), "%B %Y")

# ---- EU Vetoes: latest veto -------------------------------------------------
vetoes_csv  <- here::here("data-manual", "ms-vetoes.csv")
vetoes_df   <- read.csv(vetoes_csv, sep = ";", fileEncoding = "UTF-8-BOM") |>
  select(where(~ !(all(is.na(.)) | all(. == "")))) |>
  filter(as.Date(date_veto) > as.Date("2011-01-01"))
latest_veto <- vetoes_df |>
  arrange(desc(as.Date(date_veto))) |>
  slice(1)
veto_fact <- sprintf("%s vetoes recorded since 2011", nrow(vetoes_df))
veto_date <- format(as.Date(latest_veto$date_veto), "%B %Y")

# ---- EU Court: number of cases in the latest fully-populated year -----------
# Use the latest year that actually has cases rather than the calendar
# year — by early in a calendar year Curia hasn't yet listed any cases
# for it, so showing "0 new cases in <year>" would be misleading. The
# eu-court tracker page itself shows the calendar-year forecast.
cases <- read_with_freshness("curia_cases")
year_now <- as.integer(format(Sys.Date(), "%Y"))
latest_court_year <- cases$data |>
  filter(!is.na(case_year)) |>
  pull(case_year) |>
  max()
n_latest_year <- cases$data |>
  filter(case_year == latest_court_year) |>
  nrow()
court_fact <- sprintf("%s new cases lodged in %s",
                      format(n_latest_year, big.mark = ","),
                      latest_court_year)
court_date <- as.character(latest_court_year)

# ---- EU Law: total acts + most recent date ----------------------------------
acts <- read_with_freshness("eurlex_acts")
n_acts <- nrow(acts$data)
law_fact <- sprintf(
  "%s legal acts in Eur-Lex sector 3",
  format(n_acts, big.mark = ",")
)
law_recent_date <- acts$data |>
  filter(!is.na(date)) |>
  pull(date) |>
  as.Date() |>
  max(na.rm = TRUE)
law_date <- format(law_recent_date, "%B %Y")

# ---- EU Enlargement: current MS count + year of latest event ----------------
ms_enl <- arrow::read_parquet(here::here("data-final", "eu_member_states.parquet"))
n_current_ms <- ms_enl |> filter(is.na(exit_year)) |> nrow()
last_event <- max(c(ms_enl$accession_year, ms_enl$exit_year), na.rm = TRUE)
enl_fact <- sprintf("%s current Member States", n_current_ms)
enl_date <- as.character(last_event)

# Build one tracker card as a single un-indented HTML string. We use raw
# sprintf rather than htmltools::div() because htmltools indents nested
# children by 2 spaces and Pandoc's markdown reader treats indented HTML
# blocks as code, wrapping the rendered card content in <pre><code> on
# the live site. Single-line output dodges that entirely.
card <- function(href, title, blurb, fact, fact_date) {
  esc <- htmltools::htmlEscape
  paste0(
    '<div class="tracker-card">',
    '<a href="', esc(href), '" class="stretched-link"><h3>', esc(title), '</h3></a>',
    '<p class="tracker-blurb">', esc(blurb), '</p>',
    '<div class="tracker-fact"><strong>', esc(fact), '</strong>',
    '<span class="tracker-date"> &mdash; latest as of ', esc(fact_date), '</span></div>',
    '</div>'
  )
}
```

The European Union generates an extraordinary amount of public data — court
rulings, vetoes, legal acts, financial indicators. Most of it lives behind
unfriendly APIs and PDF archives. This site pulls those streams into
plain-language dashboards that re-render every night, so the chart you see
is the chart the data supports today.

::: {.tracker-grid}

```{r cards, echo=FALSE, results='asis'}
cat(
  card("trackers/eu-finance.html", "EU Finance",
       "Bond yields, inflation, and the ECB's key rate, plotted against the post-2008 sovereign-debt back-history.",
       fin_fact, fin_date),
  card("trackers/eu-vetoes.html", "EU Vetoes",
       "Every Member State veto on a Council file since 2010, hand-curated from the public record.",
       veto_fact, veto_date),
  card("trackers/eu-court.html", "EU Court",
       "Caseload, procedure types, rapporteur productivity, and country-of-origin maps for the CJEU.",
       court_fact, court_date),
  card("trackers/eu-law.html", "EU Law",
       "Regulations, directives, decisions, and recommendations adopted by the EU since 1952 — plus how long they took to pass.",
       law_fact, law_date),
  card("trackers/eu-enlargement.html", "EU Enlargement",
       "Membership of the EU from the Founding Six in 1957 to today, traced wave by wave (and the one withdrawal).",
       enl_fact, sprintf("last accession/exit in %s", enl_date)),
  sep = "\n"
)
```

:::

## How it works

Every tracker page is a Quarto document that loads a parquet snapshot from
`data-apis/` and renders the same set of shared `theme_mo()` charts and
`dt_mo()` tables. A GitHub Actions workflow re-fetches each upstream source
nightly at 03:17 UTC; if a fetch fails, the page falls back to the previous
good snapshot and shows a visible "data as of" badge so you always know
how stale you're looking.

The full source — fetchers, helpers, qmds, workflow — is at
[github.com/michalovadek/euverse](https://github.com/michalovadek/euverse).

## About

I'm [Michal Ovádek](https://www.ucl.ac.uk/political-science/people/dr-michal-ovadek),
a lecturer in European politics at UCL. The trackers consolidate four
separate standalone projects I'd been maintaining as one-off Rmd sites;
the goal here is one nightly render covering all four, with shared
look-and-feel and shared infrastructure.

Cite individual trackers by their printed citation block; cite the project
as a whole as: Ovádek, M. *euverse*. Accessed `r format(Sys.Date(), "%d %B %Y")`.
Available at <https://michalovadek.github.io/euverse/>.
