---
title: "euverse"
subtitle: "Public-data dashboards on European Union institutions, refreshed every night"
toc: false
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(here)
  library(htmltools)
})

# Build one tracker card as a single un-indented HTML string. We use raw
# sprintf rather than htmltools::div() because htmltools indents nested
# children by 2 spaces and Pandoc's markdown reader treats indented HTML
# blocks as code, wrapping the rendered card content in <pre><code> on
# the live site. Single-line output dodges that entirely.
card <- function(href, title, blurb) {
  esc <- htmltools::htmlEscape
  paste0(
    '<div class="tracker-card">',
    '<a href="', esc(href), '" class="stretched-link"><h3>', esc(title), '</h3></a>',
    '<p class="tracker-blurb">', esc(blurb), '</p>',
    '</div>'
  )
}
```

The European Union generates an extraordinary amount of public data:
court rulings, vetoes, legal acts, financial indicators. Most of it lives
behind unfriendly APIs and PDF archives. This site pulls those streams
into plain-language dashboards that re-render every night, so the chart
you see is the chart the data supports today.

::: {.tracker-grid}

```{r cards, echo=FALSE, results='asis'}
cat(
  card("trackers/eu-finance.html", "EU Finance",
       "Bond yields, inflation, and the ECB's key rate, plotted against the post-2008 sovereign-debt back-history."),
  card("trackers/eu-vetoes.html", "EU Vetoes",
       "Every Member State veto on a Council file since 2010, hand-curated from the public record."),
  card("trackers/eu-court.html", "EU Court",
       "Caseload, procedure types, rapporteur productivity, and country-of-origin maps for the CJEU."),
  card("trackers/eu-law.html", "EU Law",
       "Regulations, directives, decisions, and recommendations adopted by the EU since 1952, plus how long they took to pass."),
  card("trackers/eu-enlargement.html", "EU Enlargement",
       "Membership of the EU from the Founding Six in 1957 to today, traced wave by wave (and the one withdrawal)."),
  card("trackers/eu-politics.html", "EU Politics",
       "Ideological and pro/anti-EU composition of national governments since 1990, plus the EU's rise and fall as a campaign topic."),
  sep = "\n"
)
```

:::

## How it works

Every tracker page is a Quarto document that loads a parquet snapshot from
`data-apis/` and renders the same set of shared `theme_mo()` charts and
`dt_mo()` tables. A GitHub Actions workflow re-fetches each upstream source
nightly at 03:17 UTC; if a fetch fails, the page falls back to the previous
good snapshot and shows a visible "data as of" badge so you always know
how stale you're looking.

The full source (fetchers, helpers, qmds, workflow) is at
[github.com/michalovadek/euverse](https://github.com/michalovadek/euverse).

## About

I'm [Michal Ovádek](https://michalovadek.github.io/), a Lecturer in
European Institutions, Politics and Policy at UCL's Department of
Political Science. My research focuses on European integration, judicial
politics, and Central and Eastern Europe. Before UCL I held research
positions in Sweden and Belgium and worked as a political advisor in the
European Parliament.

Cite individual trackers by their printed citation block; cite the project
as a whole as: Ovádek, M. *euverse*. Accessed `r format(Sys.Date(), "%d %B %Y")`.
Available at <https://michalovadek.github.io/euverse/>.
