---
title: "euverse"
subtitle: "Public-data dashboards on European Union institutions, refreshed every night"
toc: false
page-layout: full
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(here)
  library(htmltools)
})
source(here::here("R", "freshness.R"))

# Load a few minimal facts from each tracker's snapshot so the landing
# page surfaces what the trackers actually contain instead of generic
# "EU stats here" copy. Each fact is the single line a reader would
# want to know before clicking through.

# ---- EU Finance: latest EA bond yield ---------------------------------------
yields    <- read_with_freshness("eurostat_teimf050")
ea_latest <- yields$data |>
  filter(geo == "EA") |>
  arrange(desc(date)) |>
  slice(1)
fin_fact <- sprintf(
  "Euro Area 10-year bond yield at %s%%",
  format(round(ea_latest$yield_pct, 2), nsmall = 2)
)
fin_date <- format(as.Date(ea_latest$date), "%B %Y")

# ---- EU Vetoes: latest veto -------------------------------------------------
vetoes_csv  <- here::here("data-manual", "ms-vetoes.csv")
vetoes_df   <- read.csv(vetoes_csv, sep = ";", fileEncoding = "UTF-8-BOM") |>
  select(where(~ !(all(is.na(.)) | all(. == "")))) |>
  filter(as.Date(date_veto) > as.Date("2011-01-01"))
latest_veto <- vetoes_df |>
  arrange(desc(as.Date(date_veto))) |>
  slice(1)
veto_fact <- sprintf("%s vetoes recorded since 2011", nrow(vetoes_df))
veto_date <- format(as.Date(latest_veto$date_veto), "%B %Y")

# ---- EU Court: number of cases this year + decisions this year --------------
cases <- read_with_freshness("curia_cases")
year_now <- as.integer(format(Sys.Date(), "%Y"))
n_this_year <- cases$data |>
  filter(case_year == year_now) |>
  nrow()
court_fact <- sprintf("%s new cases lodged in %s", format(n_this_year, big.mark = ","), year_now)
court_date <- format(Sys.Date(), "%B %Y")

# ---- EU Law: total acts + most recent date ----------------------------------
acts <- read_with_freshness("eurlex_acts")
n_acts <- nrow(acts$data)
law_fact <- sprintf(
  "%s legal acts in Eur-Lex sector 3",
  format(n_acts, big.mark = ",")
)
law_recent_date <- acts$data |>
  filter(!is.na(date)) |>
  pull(date) |>
  as.Date() |>
  max(na.rm = TRUE)
law_date <- format(law_recent_date, "%B %Y")

# Build one tracker card.
card <- function(href, title, blurb, fact, fact_date) {
  htmltools::div(
    class = "tracker-card",
    htmltools::tags$a(
      href = href, class = "stretched-link",
      htmltools::h3(title)
    ),
    htmltools::p(blurb, class = "tracker-blurb"),
    htmltools::div(
      class = "tracker-fact",
      htmltools::strong(fact),
      htmltools::tags$span(class = "tracker-date",
                           paste0(" — latest as of ", fact_date))
    )
  )
}
```

The European Union generates an extraordinary amount of public data — court
rulings, vetoes, legal acts, financial indicators. Most of it lives behind
unfriendly APIs and PDF archives. This site pulls four such streams into
plain-language dashboards that re-render every night, so the chart you see
is the chart the data supports today.

::: {.tracker-grid}

```{r cards, echo=FALSE, results='asis'}
htmltools::tagList(
  card(
    "trackers/eu-finance.html",
    "EU Finance",
    "Bond yields, inflation, and the ECB's key rate, plotted against the post-2008 sovereign-debt back-history.",
    fin_fact, fin_date
  ),
  card(
    "trackers/eu-vetoes.html",
    "EU Vetoes",
    "Every Member State veto on a Council file since 2010, hand-curated from the public record.",
    veto_fact, veto_date
  ),
  card(
    "trackers/eu-court.html",
    "EU Court",
    "Caseload, procedure types, rapporteur productivity, and country-of-origin maps for the CJEU.",
    court_fact, court_date
  ),
  card(
    "trackers/eu-law.html",
    "EU Law",
    "Regulations, directives, decisions, and recommendations adopted by the EU since 1952 — plus how long they took to pass.",
    law_fact, law_date
  )
) |> htmltools::tagList() |> as.character() |> cat()
```

:::

## How it works

Every tracker page is a Quarto document that loads a parquet snapshot from
`data-apis/` and renders the same set of shared `theme_mo()` charts and
`dt_mo()` tables. A GitHub Actions workflow re-fetches each upstream source
nightly at 03:17 UTC; if a fetch fails, the page falls back to the previous
good snapshot and shows a visible "data as of" badge so you always know
how stale you're looking.

The full source — fetchers, helpers, qmds, workflow — is at
[github.com/michalovadek/euverse](https://github.com/michalovadek/euverse).

## About

I'm [Michal Ovádek](https://www.ucl.ac.uk/political-science/people/dr-michal-ovadek),
a lecturer in European politics at UCL. The trackers consolidate four
separate standalone projects I'd been maintaining as one-off Rmd sites;
the goal here is one nightly render covering all four, with shared
look-and-feel and shared infrastructure.

Cite individual trackers by their printed citation block; cite the project
as a whole as: Ovádek, M. *euverse*. Accessed `r format(Sys.Date(), "%d %B %Y")`.
Available at <https://michalovadek.github.io/euverse/>.
