---
title: "Judicial Proceedings in the European Union"
subtitle: "An automatically updated overview of the EU's judicial activity"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(forcats)
  library(purrr)
  library(ggplot2)
  library(ggiraph)
  library(here)
  library(maps)      # provides map data behind ggplot2::map_data
  library(mapproj)   # provides projections used by coord_map()
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load snapshots once at the top.
cc  <- read_with_freshness("curia_cases")
ed  <- read_with_freshness("eurlex_court_decisions")
ms  <- arrow::read_parquet(here::here("data-final", "eu_member_states.parquet"))

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))

curia_cases <- cc$data
decisions   <- ed$data

# Year x Member State panel from the shared master.
ms_panel <- tibble(year = 1952:year_now) |>
  tidyr::crossing(ms) |>
  filter(year >= accession_year,
         is.na(exit_year) | year < exit_year) |>
  select(year, country, iso2)

# Yearly count of MS for parity with the eu-law port's reuse pattern.
n_ms_by_year <- ms_panel |> count(year, name = "n_ms")

# Inline-stats: caseload totals.
n_cases       <- curia_cases |> filter(case_status != "Transferred" | is.na(case_status)) |> summarise(n = n_distinct(case_id)) |> pull(n)
n_decisions   <- curia_cases |> filter(!is.na(ecli)) |> summarise(n = n_distinct(ecli)) |> pull(n)
n_pending_cj  <- curia_cases |> filter(case_status == "Pending", court == "CJ") |> summarise(n = n_distinct(case_id)) |> pull(n)
n_pending_gc  <- curia_cases |> filter(case_status == "Pending", court == "GC") |> summarise(n = n_distinct(case_id)) |> pull(n)

# Court palette mapped to mo_palette$categorical[1:3] (replaces old bespoke pink/navy/teal).
court_palette <- tibble(
  court      = factor(c("CJ", "GC", "CST"), levels = c("CJ", "GC", "CST")),
  court_long = factor(c("Court of Justice", "General Court", "Civil Service Tribunal"),
                      levels = c("Court of Justice", "General Court", "Civil Service Tribunal")),
  color      = mo_palette$categorical[1:3]
)
```

The Court of Justice of the European Union (CJEU) is the principal
judicial institution of the EU. It consists of two courts: the
higher-instance Court of Justice (CJ) and the lower-instance General
Court (GC). There used to be a third court, the Civil Service Tribunal
(CST), which was abolished and whose jurisdiction was transferred to
the GC.

This page gives a real-time overview of the EU courts' output and
engagement with national courts through the preliminary ruling
procedure. It relies on data collected via the
[eurlex](https://github.com/michalovadek/eurlex) package for R.^[Because
neither [Eur-Lex](https://eur-lex.europa.eu/homepage.html) nor
[Curia](https://curia.europa.eu/juris/) contains the complete record of
all court proceedings and documents, the data below should be viewed as
a reasonable approximation of the true state of the world. The
[IUROPA](https://iuropa.pol.gu.se/) project is attempting to create
more comprehensive datasets.]

As of `r format(today_d, "%d %B %Y")`, EU courts have delivered
`r mono(fmt_n(n_decisions))` decisions in about
`r mono(fmt_n(n_cases))` cases submitted to them since 1953.
According to public records, `r mono(fmt_n(n_pending_cj))` cases are
currently pending before the Court of Justice and
`r mono(fmt_n(n_pending_gc))` before the General Court.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(cc$meta, ed$meta))
```

```{r recent-cj-cases, echo=FALSE}
curia_cases |>
  filter(court == "CJ") |>
  arrange(desc(case_year), desc(case_number)) |>
  slice(1:50) |>
  transmute(
    `Case ID` = case_id,
    Info      = case_info
  ) |>
  dt_mo()
```

# Number of cases

Even more than any other branch of the EU, the CJEU's caseload changed
dramatically over the course of European integration. We can visualise
the annual caseload of the three EU courts since 1953.

```{r ncasesyearcourt, echo=FALSE, fig.height=8, fig.width=9}
n_cases_year_court <- curia_cases |>
  filter(!is.na(court), !is.na(case_year)) |>
  summarise(n = n_distinct(case_id), .by = c(court, case_year)) |>
  right_join(tidyr::expand(curia_cases |> filter(!is.na(court), !is.na(case_year)),
                           court, case_year), by = c("court", "case_year")) |>
  mutate(n = ifelse(is.na(n), 0L, n)) |>
  left_join(court_palette, by = "court") |>
  group_by(court) |>
  mutate(global_mean = mean(n)) |>
  ungroup() |>
  mutate(tooltip = paste0("N = ", n, " (", case_year, ")"),
         data_id = paste0(court, case_year))

p_caseload <- ggplot(n_cases_year_court,
                     aes(x = case_year, y = n, fill = color)) +
  geom_hline(aes(yintercept = global_mean, colour = color),
             lty = 2, alpha = 0.4, linewidth = 0.4) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = data_id),
                       colour = "grey96") +
  scale_fill_identity() +
  scale_colour_identity() +
  scale_x_continuous(breaks = seq(1950, year_now + 1, by = 10)) +
  scale_y_continuous(breaks = int_breaks()) +
  facet_wrap(~ court_long, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL,
       title = "Number of cases received by EU courts",
       subtitle = "Aggregated by year and court",
       caption = "Dashed line shows the mean") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_caseload, width_svg = 9, height_svg = 8)
```

The increase in the CJ's and GC's caseloads was mostly gradual.
Interestingly, the outsourcing of the civil service agenda to the CST
did not prevent the GC's caseload from rising during its period of
operation.

# Forecasting

As most other organisations, EU courts operate under a budget
constraint while facing some uncertainty about their future workload.
With information on the number of cases received in the past in hand,
we can build a simple autoregressive model that predicts this year's
and next year's CJ caseload.

The model uses lags 1, 2 and 3 of CJ caseload plus a 3-year lag of GC
caseload (since CJ caseload is linked to GC's through appeal
procedures):

$$
Y_{t}^{CJ} = \beta_0 + \beta_1 Y_{t-1}^{CJ} + \beta_2 Y_{t-2}^{CJ} + \beta_3 Y_{t-3}^{CJ} + \beta_4 Y_{t-3}^{GC} + \epsilon_t
$$

The parameters are estimated via Gaussian or Poisson regression.^[Because
we are working with count data, Poisson regression should be in principle
more appropriate.]

```{r forecast, echo=FALSE, fig.height=4, fig.width=9}
dat_reg <- n_cases_year_court |>
  filter(case_year > 1994, court != "CST") |>
  arrange(case_year) |>
  group_by(court) |>
  mutate(lag1 = lag(n), lag2 = lag(n, 2), lag3 = lag(n, 3)) |>
  ungroup() |>
  filter(case_year > 2000) |>
  select(court, case_year, n, lag1, lag2, lag3) |>
  tidyr::pivot_wider(id_cols = case_year, names_from = court,
                     values_from = c(n, lag1, lag2, lag3))

out_ols <- lm(data = dat_reg |> filter(case_year < year_now),
              formula = n_CJ ~ lag1_CJ + lag2_CJ + lag3_CJ + lag3_GC)
out_poi <- glm(data = dat_reg |> filter(case_year < year_now),
               family = poisson(),
               formula = n_CJ ~ lag1_CJ + lag2_CJ + lag3_CJ + lag3_GC)

yhat_ols <- as.integer(round(predict(out_ols, dat_reg[nrow(dat_reg), ], type = "response"), 0))
yhat_poi <- as.integer(round(predict(out_poi, dat_reg[nrow(dat_reg), ], type = "response"), 0))

# t+1 prediction (use averaged prediction as new lag1_CJ)
newdat_t1 <- tibble(
  case_year = year_now + 1,
  lag1_CJ   = (yhat_ols + yhat_poi) / 2,
  lag2_CJ   = dat_reg$lag1_CJ[nrow(dat_reg)],
  lag3_CJ   = dat_reg$lag2_CJ[nrow(dat_reg)],
  lag3_GC   = dat_reg$lag2_GC[nrow(dat_reg)]
)
yhat_ols_t1 <- as.integer(round(predict(out_ols, newdat_t1, type = "response"), 0))

cj_color <- court_palette$color[court_palette$court == "CJ"]

# Plot: actual bars + 2 dashed-outline prediction bars.
#
# Curia's case-list scrape lags upstream by several months for the
# current year, so n_cases_year_court is often missing a year_now row
# despite cases being publicly registered. When curia is behind, fall
# back to eurlex_court_decisions: every decision celex encodes a case
# year + court + case_number ("62026CJ0024" = case 24 of 2026, CJ).
# CJEU case numbers are sequential, so the MAX case_number observed
# in year_now decisions is a tight lower bound on cases lodged so far.
# This is what makes the actual bar appear inside the dashed year_now
# prediction bar instead of nothing showing.
plot_dat <- n_cases_year_court |>
  filter(case_year > year_now - 7, court == "CJ")

if (!any(plot_dat$case_year == year_now)) {
  cj_partial <- decisions |>
    filter(clx_court == "C", clx_year == year_now, !is.na(celex)) |>
    mutate(case_num = suppressWarnings(as.integer(
      stringr::str_extract(celex, "(?<=[A-Z])[0-9]+")))) |>
    filter(!is.na(case_num)) |>
    summarise(max_case = max(case_num, na.rm = TRUE)) |>
    pull(max_case)

  if (length(cj_partial) == 1L && is.finite(cj_partial) && cj_partial > 0) {
    plot_dat <- dplyr::bind_rows(plot_dat, tibble::tibble(
      court       = "CJ",
      case_year   = year_now,
      n           = as.integer(cj_partial),
      color       = cj_color,
      global_mean = NA_real_
    ))
  }
}

plot_dat <- plot_dat |>
  mutate(tooltip = ifelse(
           case_year == year_now,
           paste0("≥ ", n, " cases lodged so far in ", year_now,
                  " (lower bound from EUR-Lex case numbers; Curia case list still behind)"),
           paste0("N = ", n)
         ),
         data_id = paste0("actual", case_year))

p_forecast <- ggplot(plot_dat, aes(x = case_year, y = n)) +
  # Forecast bars BEHIND the actual bars. alpha 0.4 + linewidth 0.6
  # makes the predicted bars clearly visible (was 0.15, barely showed).
  geom_col_interactive(
    data = tibble(case_year = year_now, n = yhat_ols,
                  tooltip = paste0("Forecast for ", year_now, " = ", yhat_ols),
                  data_id = paste0("pred", year_now)),
    aes(tooltip = tooltip, data_id = data_id),
    colour = cj_color, fill = cj_color, alpha = 0.4, lty = 2, linewidth = 0.6
  ) +
  geom_col_interactive(
    data = tibble(case_year = year_now + 1, n = yhat_ols_t1,
                  tooltip = paste0("Forecast for ", year_now + 1, " = ", yhat_ols_t1),
                  data_id = paste0("pred", year_now + 1)),
    aes(tooltip = tooltip, data_id = data_id),
    colour = cj_color, fill = cj_color, alpha = 0.4, lty = 2, linewidth = 0.6
  ) +
  geom_col_interactive(aes(fill = color, tooltip = tooltip, data_id = data_id)) +
  scale_fill_identity() +
  scale_x_continuous(breaks = (year_now - 7):(year_now + 1)) +
  scale_y_continuous(breaks = int_breaks()) +
  labs(x = NULL, y = NULL,
       title = "Number of cases received by the Court of Justice",
       subtitle = "Forecast for the current and next year",
       caption = "Predicted years shown with dashed border + lighter fill") +
  theme_mo()

girafe(ggobj = p_forecast, width_svg = 9, height_svg = 4)
```

```{r forecast-table, echo=FALSE}
suppressPackageStartupMessages(library(modelsummary))
modelsummary::modelsummary(
  list("OLS" = out_ols, "Poisson" = out_poi),
  statistic  = "conf.int",
  conf_level = 0.95,
  stars      = TRUE,
  title      = "Autoregressive model of CJ yearly caseload"
)
```

Both Gaussian and Poisson regression yield comparable predictions for
`r year_now` (`r yhat_ols` versus `r yhat_poi` cases).

# Decisions

The number of cases received is not identical to the number of judicial
decisions produced. Some cases are withdrawn or otherwise discontinued,
others are joined and decided together. The EU judicial system
distinguishes between different types of public decisions, of which
judgments and orders are the two most common.

```{r decisiontype, echo=FALSE, fig.height=5, fig.width=9}
n_dec_type_proc <- decisions |>
  filter(decision != "Opinion",
         procedure_class != "Other",
         clx_court == "C") |>
  mutate(decision_year = as.integer(str_sub(date, 1, 4))) |>
  count(decision_year, decision, procedure_class) |>
  tidyr::drop_na() |>
  mutate(tooltip = paste0(decision, "s = ", n, " (", decision_year, ")"),
         data_id = paste0("C", decision_year, procedure_class, decision))

p_dec_type <- ggplot(n_dec_type_proc,
                     aes(x = decision_year, y = n, fill = decision)) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = data_id),
                       colour = "grey96", show.legend = TRUE) +
  scale_fill_manual(values = mo_palette$categorical[c(1, 2)]) +
  scale_x_continuous(breaks = seq(1950, year_now + 1, by = 10)) +
  facet_wrap(~ fct_infreq(procedure_class), dir = "v", scales = "free_y") +
  labs(x = NULL, y = NULL,
       title = "Decisions handed down by the Court of Justice",
       subtitle = "By year and procedure type",
       fill = NULL) +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_dec_type, width_svg = 9, height_svg = 5)
```

The four main procedure types are preliminary rulings (Art 267 TFEU),
annulment procedures (Art 263), infringement proceedings (Art 258) and
staff cases. The CJ has been handing down a seemingly ever-increasing
number of preliminary rulings on questions from national courts.

```{r formationproportion, echo=FALSE, fig.height=4, fig.width=9}
form_dat <- decisions |>
  filter(clx_court == "C") |>
  mutate(decision_year = as.integer(str_sub(date, 1, 4))) |>
  count(decision_year, formation_simple) |>
  filter(decision_year > 2004) |>
  group_by(decision_year) |>
  mutate(prop = n / sum(n)) |>
  ungroup() |>
  mutate(formation_simple = factor(formation_simple,
                                   levels = c("Full Court", "Grand Chamber", "Panel", "Single Judge")),
         tooltip = paste0(formation_simple, " = ", round(prop, 3) * 100, "% (", decision_year, ")"),
         data_id = paste0(decision_year, formation_simple))

p_formation <- ggplot(form_dat,
                      aes(x = decision_year, y = prop, fill = formation_simple)) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = data_id),
                       colour = "grey96", show.legend = TRUE) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = mo_palette$categorical[c(1, 2, 3, 4)]) +
  scale_x_continuous(breaks = seq(2005, year_now + 1, by = 3)) +
  labs(x = NULL, y = NULL,
       title = "Proportion of CJ decisions by formation",
       fill = NULL) +
  theme_mo()

girafe(ggobj = p_formation, width_svg = 9, height_svg = 4)
```

Decisions can be further classified by the deciding court formation.
The most important cases at the CJ are decided by more than ten judges
in the Grand Chamber; on special occasions, the CJ even sits as the
Full Court. Most cases are decided by a panel of five judges.

```{r rapporteurs, echo=FALSE, fig.height=9, fig.width=9}
top_rapps <- decisions |>
  filter(clx_court %in% c("C", "T"),
         procedure_class != "Other",
         decision == "Judgment",
         !is.na(rapporteur), rapporteur != "") |>
  mutate(court_long = ifelse(clx_court == "C", "Court of Justice", "General Court")) |>
  count(court_long, rapporteur) |>
  group_by(court_long) |>
  slice_max(n, n = 20) |>
  ungroup()

rapp_dat <- decisions |>
  filter(clx_court %in% c("C", "T"),
         procedure_class != "Other",
         decision == "Judgment",
         !is.na(rapporteur), rapporteur != "") |>
  mutate(court_long = ifelse(clx_court == "C", "Court of Justice", "General Court")) |>
  semi_join(top_rapps, by = c("court_long", "rapporteur")) |>
  count(court_long, rapporteur, procedure_class) |>
  mutate(sum = sum(n), .by = c(court_long, rapporteur),
         tooltip = paste0(procedure_class, " = ", n, " (", rapporteur, ")"),
         data_id = paste0(court_long, rapporteur, procedure_class))

p_rapp <- ggplot(rapp_dat,
                 aes(x = n, y = reorder(rapporteur, sum), fill = procedure_class)) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = data_id),
                       colour = "grey96", position = "stack", show.legend = TRUE) +
  scale_fill_manual(values = mo_palette$categorical[c(1, 2, 5, 3)]) +
  # Stack panels vertically: with 17+-char surnames ("Silva de Lapuerta",
  # "Moitinho de Almeida") the side-by-side layout crams labels into ~78 px
  # of y-axis space per panel. ncol = 1 gives each panel the full chart
  # width and lets ggplot auto-allocate enough label space.
  facet_wrap(~ court_long, ncol = 1, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "Most frequent judge-rapporteurs",
       subtitle = "Judgments in the four main procedure types only",
       fill = NULL) +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_rapp, width_svg = 9, height_svg = 9)
```

Each decision is assigned a judge-rapporteur. This judge has the
primary responsibility for drafting the decision. The main predictor of
how many decisions a judge writes is the length of their tenure.

# Preliminary references

Historically the most important subset of CJEU cases comes through the
preliminary ruling procedure (Art 267 TFEU). The most well-known CJ
decisions, such as _Costa v ENEL_ or _Cassis de Dijon_, originated in references
from national courts.

```{r prelrefsmap, echo=FALSE, fig.height=6, fig.width=9}
references <- decisions |>
  filter(str_detect(procedure_class, "Preliminary")) |>
  tidyr::separate_rows(origin, sep = "~~~") |>
  filter(!is.na(origin), origin != "")

n_refs_country <- references |>
  filter(!is.na(clx_year)) |>
  arrange(date) |>
  group_by(origin) |>
  summarise(year_first = min(clx_year),
            n_refs     = n(), .groups = "drop") |>
  mutate(tooltip = paste0(origin, ": ", n_refs, " references. First case referred in ", year_first, "."))

# ggplot2::map_data + Czech Republic / UK name normalisation
refs_map_dat <- ggplot2::map_data("world",
                                  xlim = c(-10.66, 51.55),
                                  ylim = c(34.5, 71.05)) |>
  mutate(region = ifelse(region == "Czech Republic", "Czechia", region),
         region = ifelse(region == "UK", "United Kingdom", region))

# Country centroids for label overlay. Hardcoded because the map data
# is polygon outlines, not centroids, and the small countries (MT,
# CY, LU) are otherwise unfindable on a Europe-scale map. Each row
# carries the ISO-2 code shown over the country and the full name for
# the tooltip + map join key.
eu_centroids <- tibble::tribble(
  ~origin,          ~iso2,  ~lon,   ~lat,
  "Austria",         "AT",  14.5,   47.5,
  "Belgium",         "BE",   4.5,   50.6,
  "Bulgaria",        "BG",  25.5,   42.7,
  "Croatia",         "HR",  15.8,   45.3,
  "Cyprus",          "CY",  33.2,   35.0,
  "Czechia",         "CZ",  15.5,   49.7,
  "Denmark",         "DK",   9.5,   56.0,
  "Estonia",         "EE",  25.5,   58.6,
  "Finland",         "FI",  26.0,   64.5,
  "France",          "FR",   2.5,   46.5,
  "Germany",         "DE",  10.5,   51.0,
  "Greece",          "GR",  22.0,   39.0,
  "Hungary",         "HU",  19.5,   47.0,
  "Ireland",         "IE",  -8.0,   53.0,
  "Italy",           "IT",  12.5,   42.5,
  "Latvia",          "LV",  24.6,   56.9,
  "Lithuania",       "LT",  23.9,   55.3,
  "Luxembourg",      "LU",   6.1,   49.8,
  "Malta",           "MT",  14.5,   35.9,
  "Netherlands",     "NL",   5.5,   52.2,
  "Poland",          "PL",  19.5,   52.0,
  "Portugal",        "PT",  -8.0,   39.5,
  "Romania",         "RO",  25.0,   45.8,
  "Slovakia",        "SK",  19.5,   48.7,
  "Slovenia",        "SI",  14.5,   46.1,
  "Spain",           "ES",  -3.7,   40.0,
  "Sweden",          "SE",  16.0,   62.0,
  "United Kingdom",  "GB",  -2.0,   53.0
) |>
  left_join(n_refs_country |> select(origin, n_refs), by = "origin") |>
  mutate(label_tooltip = paste0(origin, ": ",
                                ifelse(is.na(n_refs), "0", format(n_refs, big.mark = ",")),
                                " preliminary references"))

p_map <- ggplot(n_refs_country) +
  geom_map_interactive(
    aes(map_id = origin, fill = n_refs, tooltip = tooltip, data_id = origin),
    map = refs_map_dat, colour = "grey60", linewidth = 0.2, show.legend = FALSE
  ) +
  # Country code labels at each centroid so small states (MT, CY, LU)
  # are findable. White text on dark border so it stays readable across
  # the viridis fill range.
  ggiraph::geom_text_interactive(
    data = eu_centroids,
    aes(x = lon, y = lat, label = iso2,
        tooltip = label_tooltip, data_id = origin),
    colour = "#ffffff", fontface = "bold", size = 2.8,
    inherit.aes = FALSE
  ) +
  expand_limits(x = c(-13.66, 38.55), y = c(34.5, 71.05)) +
  # Albers conic projection via coord_map (needs mapproj, pinned in
  # renv.lock). Standard parallels at 40 and 60 °N give a nice
  # equal-area look for Europe, much better than the default
  # unprojected stretch.
  coord_map("albers", lat0 = 40, lat1 = 60) +
  scale_fill_gradientn(colours = mo_palette$sequential(9), na.value = "grey85") +
  labs(x = NULL, y = NULL,
       title = "Preliminary references by country of origin",
       subtitle = "Total number of closed cases referred by national courts") +
  theme_mo() +
  theme(panel.grid = element_blank(),
        axis.text  = element_blank())

girafe(ggobj = p_map, width_svg = 9, height_svg = 6)
```

References from national courts are very unevenly distributed
geographically. By far the highest number of references has come from
Germany. There is a relationship between length of EU membership and
total references.

```{r prelreftime, echo=FALSE, fig.height=7, fig.width=9}
# Year x country panel of references with zeros filled in.
ms_active_years <- ms_panel |> rename(origin = country)
n_ref_year <- references |>
  rename(year = clx_year) |>
  group_by(origin, year) |>
  summarise(n_refs = n(), .groups = "drop")

n_ref_year_all <- ms_active_years |>
  filter(year > 1960) |>
  left_join(n_ref_year, by = c("origin", "year")) |>
  mutate(n_refs = ifelse(is.na(n_refs), 0L, n_refs),
         tooltip = paste0("N = ", n_refs, " (", year, ")"),
         data_id = paste0(origin, year))

p_refs_time <- ggplot(n_ref_year_all, aes(x = year, y = n_refs)) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = data_id),
                         colour = mo_palette$categorical[2], alpha = 0.4, size = 0.9) +
  geom_smooth(colour = mo_palette$categorical[6], fill = mo_palette$categorical[8],
              alpha = 0.15, linewidth = 0.6) +
  scale_y_continuous(limits = c(0, NA),
                     breaks = function(x) unique(floor(pretty(seq(0, (max(x) + 1) * 1.1))))) +
  scale_x_continuous(breaks = scales::pretty_breaks(3)) +
  facet_wrap(~ origin, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "Preliminary references over time",
       subtitle = "By country of origin (free y-scale per panel)") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold", size = 6.5),
        axis.text  = element_text(size = 5))

girafe(ggobj = p_refs_time, width_svg = 9, height_svg = 7)
```

Referral activity has increased over time in most countries as the CJEU
and its preliminary-ruling procedure gained prominence. Note that the
data here comes from _closed_ cases only, so many recent referrals are
still pending.

# Cite

Ovádek, M. *Judicial Proceedings in the European Union*. Accessed
`r format(Sys.Date(), "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-court.html>.
