---
title: "Enlargement of the European Union"
subtitle: "From six to twenty-seven (and back to twenty-seven): the EU's membership history"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggiraph)
  library(here)
  library(stringr)
  library(forcats)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

ms_csv  <- here::here("data-manual", "eu-member-states.csv")
ms_meta <- manual_meta(ms_csv, source_label = "data-manual/eu-member-states.csv (hand-curated)")

ms <- arrow::read_parquet(here::here("data-final", "eu_member_states.parquet"))
ms_meta$rows <- nrow(ms)

cand_csv <- here::here("data-manual", "eu-candidate-states.csv")
candidates <- read.csv(cand_csv, na.strings = "") |>
  mutate(candidate_year     = as.integer(candidate_year),
         negotiations_year  = as.integer(negotiations_year))

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))

# Membership-count panel: each row a (year, country) pair with TRUE if
# that country was an EU MS in that year. Aggregate to year totals
# downstream; reuse for the waves chart too.
ms_long <- tibble(year = 1952:year_now) |>
  tidyr::crossing(ms) |>
  filter(year >= accession_year,
         is.na(exit_year) | year < exit_year)

count_by_year <- ms_long |> count(year, name = "n_ms")

# Accession-wave bucketing for the area chart fill.
# Six founding states 1952, then the 1973, 1981, 1986, 1995, 2004,
# 2007, 2013 enlargements. UK leaves the EU in 2020 (Brexit).
ms_wave <- ms |>
  mutate(wave = case_when(
    accession_year == 1952 ~ "1952 — Founding 6 (BE, DE, FR, IT, LU, NL)",
    accession_year == 1973 ~ "1973 — North-Western (DK, IE, UK)",
    accession_year == 1981 ~ "1981 — Greece",
    accession_year == 1986 ~ "1986 — Iberia (PT, ES)",
    accession_year == 1995 ~ "1995 — EFTA (AT, FI, SE)",
    accession_year == 2004 ~ "2004 — Eastern Big Bang (10 states)",
    accession_year == 2007 ~ "2007 — Bulgaria + Romania",
    accession_year == 2013 ~ "2013 — Croatia",
    TRUE                    ~ NA_character_
  )) |>
  mutate(wave = factor(wave, levels = c(
    "1952 — Founding 6 (BE, DE, FR, IT, LU, NL)",
    "1973 — North-Western (DK, IE, UK)",
    "1981 — Greece",
    "1986 — Iberia (PT, ES)",
    "1995 — EFTA (AT, FI, SE)",
    "2004 — Eastern Big Bang (10 states)",
    "2007 — Bulgaria + Romania",
    "2013 — Croatia"
  )))

# Per-year membership tagged with each MS's wave - needed for the
# stacked area chart.
ms_long_wave <- ms_long |>
  left_join(ms_wave |> select(iso2, wave), by = "iso2") |>
  count(year, wave, name = "n_ms")
```

The European Economic Community started life in 1957 with six members.
By 2020 it had become the European Union of 27 — having passed through
`r mono(length(unique(ms_wave$wave)))` distinct enlargement waves and one
withdrawal. This page traces that history from the source data the rest
of the euverse trackers depend on: the small hand-curated master file
listing every Member State's accession year and (where applicable) exit
year.

```{r freshness, results='asis', echo=FALSE}
freshness_badge(ms_meta)
```

# Member State count over time

```{r ms-count, echo=FALSE, fig.height=4, fig.width=9}
ggplot(ms_long_wave, aes(x = year, y = n_ms, fill = wave)) +
  geom_area(colour = "white", linewidth = 0.15) +
  scale_fill_manual(values = mo_palette$categorical) +
  scale_x_continuous(expand = c(0, 0), breaks = seq(1950, year_now, 10)) +
  scale_y_continuous(expand = c(0, 0), breaks = c(6, 9, 10, 12, 15, 25, 27, 28),
                     limits = c(0, 30)) +
  labs(x = NULL, y = "Number of Member States",
       title = "EU membership over time",
       subtitle = "Stacked by accession wave; the dip in 2020 is Brexit",
       fill = NULL) +
  theme_mo() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(nrow = 3, byrow = TRUE))
```

The 2004 "Big Bang" enlargement remains the largest single accession in
EU history: ten countries — most of post-Soviet Central and Eastern
Europe plus Malta and Cyprus — joined on the same day. The chart's only
dip is **2020**, when the United Kingdom became the first (and so far
only) Member State to leave.

# Accession timeline

```{r waves-timeline, echo=FALSE, fig.height=6, fig.width=9}
ms_timeline <- ms_wave |>
  mutate(end_year = ifelse(is.na(exit_year), year_now, exit_year)) |>
  arrange(accession_year, country) |>
  mutate(country = factor(country, levels = country))

p_timeline <- ggplot(ms_timeline,
                     aes(y = country, x = accession_year, xend = end_year,
                         colour = wave,
                         tooltip = paste0(country, " (", iso2, "): ",
                                          accession_year,
                                          ifelse(is.na(exit_year), "—now",
                                                 paste0("—", exit_year))),
                         data_id = iso2)) +
  geom_segment_interactive(aes(yend = country), linewidth = 4, lineend = "round") +
  geom_point_interactive(aes(x = accession_year), size = 2.6) +
  scale_colour_manual(values = mo_palette$categorical) +
  scale_x_continuous(breaks = seq(1950, year_now, 10), expand = c(0.02, 0.02)) +
  labs(x = NULL, y = NULL,
       title = "When each Member State joined (and, for the UK, left)",
       subtitle = "Each bar runs from the accession year to today; UK's bar ends in 2020",
       colour = NULL) +
  theme_mo() +
  theme(panel.grid.major.y = element_blank(),
        legend.position    = "bottom",
        legend.text        = element_text(size = 8)) +
  guides(colour = guide_legend(nrow = 3, byrow = TRUE))

girafe(ggobj = p_timeline, width_svg = 9, height_svg = 6)
```

# Member State table

```{r ms-table, echo=FALSE}
ms_wave |>
  arrange(accession_year, country) |>
  transmute(
    Country       = country,
    `ISO-2`       = iso2,
    `Joined`      = accession_year,
    `Left`        = ifelse(is.na(exit_year), "—", as.character(exit_year)),
    `Years as MS` = ifelse(is.na(exit_year),
                           year_now - accession_year,
                           exit_year - accession_year),
    Wave          = stringr::str_remove(as.character(wave), " — .*$")
  ) |>
  dt_mo()
```

# Candidate and potential-candidate countries

The chart and table above only show **current and historical Member
States**. As of `r format(today_d, "%B %Y")`,
`r mono(sum(candidates$status == "candidate"))` countries hold formal
candidate status and
`r mono(sum(candidates$status == "potential candidate"))` hold
potential-candidate status. Together they describe the EU's current
enlargement pipeline.

```{r candidates-timeline, echo=FALSE, fig.height=4.5, fig.width=9}
# Chart shows only countries with a candidate_year (i.e. excludes
# Kosovo, which has applied but has no formal candidate status yet;
# Kosovo still appears in the table below). Sort by accession-status
# year, descending — newest aspirants at the top.
cand_plot <- candidates |>
  filter(!is.na(candidate_year)) |>
  mutate(end   = year_now,
         # Short name for the y-axis label so "Bosnia and Herzegovina"
         # doesn't crowd the left margin. Full country name and status
         # date live in the tooltip already.
         label = case_when(
           country == "Bosnia and Herzegovina" ~ "Bosnia & Herz. (BA)",
           country == "North Macedonia"        ~ "N. Macedonia (MK)",
           TRUE                                ~ paste0(country, " (", iso2, ")")
         )) |>
  arrange(desc(candidate_year), country) |>
  mutate(label = factor(label, levels = label))

p_cand <- ggplot(cand_plot,
                 aes(y = label,
                     tooltip = paste0(country, ": ", status,
                                      " since ", candidate_year,
                                      ifelse(is.na(negotiations_year), "",
                                             paste0(", negotiations from ",
                                                    negotiations_year))),
                     data_id = iso2)) +
  geom_segment_interactive(aes(x = candidate_year, xend = end, yend = label,
                               colour = status),
                           linewidth = 4, lineend = "round") +
  geom_point_interactive(aes(x = negotiations_year),
                         size = 3, shape = 21, fill = "white",
                         colour = "grey30",
                         na.rm = TRUE) +
  scale_colour_manual(values = c(candidate = mo_palette$categorical[5],
                                 `potential candidate` = mo_palette$categorical[8])) +
  scale_x_continuous(breaks = seq(1995, year_now, 5), expand = c(0.02, 0.02)) +
  labs(x = NULL, y = NULL,
       title = "Current EU enlargement pipeline",
       subtitle = "Bar runs from candidate-status year; ring marks year accession negotiations opened",
       colour = NULL,
       caption = "Türkiye: negotiations opened 2005 but effectively frozen since 2018. Kosovo applied 2022 but is not yet a candidate (table below).") +
  theme_mo() +
  theme(panel.grid.major.y = element_blank())

girafe(ggobj = p_cand, width_svg = 9, height_svg = 4.5)
```

```{r candidates-table, echo=FALSE}
candidates |>
  arrange(country) |>
  transmute(
    Country                = country,
    `ISO-2`                = iso2,
    Status                 = status,
    `Candidate since`      = ifelse(is.na(candidate_year), "—",
                                    as.character(candidate_year)),
    `Negotiations since`   = ifelse(is.na(negotiations_year), "—",
                                    as.character(negotiations_year))
  ) |>
  dt_mo()
```

# About this page

Source: [`data-manual/eu-member-states.csv`](https://github.com/michalovadek/euverse/blob/main/data-manual/eu-member-states.csv),
a hand-curated list of every state that has ever held EU membership,
with its accession year and (where applicable) exit year. Built into
[`data-final/eu_member_states.parquet`](https://github.com/michalovadek/euverse/blob/main/data-final/eu_member_states.parquet)
by [`R/build_eu_member_states.R`](https://github.com/michalovadek/euverse/blob/main/R/build_eu_member_states.R).
The same parquet is consumed by the
[EU Court](eu-court.html) and [EU Law](eu-law.html) trackers for their
member-state-count time-series.

Cite as: Ovádek, M. *Enlargement of the European Union*. Accessed
`r format(Sys.Date(), "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-enlargement.html>.
