---
title: "Borrowing and Yields in the European Union"
subtitle: "An automatically updated overview of the EU's finances"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggiraph)
  library(ggforce)
  library(dplyr)
  library(lubridate)
  library(stringr)
  library(here)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load all four snapshots once at the top of the page.
yc  <- read_with_freshness("ecb_yield_curve")
eus <- read_with_freshness("eurostat_teimf050")
mrr <- read_with_freshness("ecb_mrr")
hic <- read_with_freshness("ecb_hicp")

today_d <- Sys.Date()
```

The European Union is not only a single market but also an economic and
monetary union. Most members share the euro, which entails a supranational
monetary policy run by the European Central Bank (ECB) together with Euro
Area central banks. This page provides an automatically updated overview of
several indicators central to that policy, sourced from the
[ECB Data Portal API](https://data.ecb.europa.eu/help/api/) and Eurostat.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(yc$meta, eus$meta, mrr$meta, hic$meta))
```

# Euro Area government bond yields

Government borrowing is normally carried out through bond sales. Bonds entitle
the buyer to an interest rate (coupon) and the return of the principal at
maturity. Bonds trade on secondary markets and their yields fluctuate
accordingly.

```{r euro-area-yields, echo=FALSE, fig.height=4.5, fig.width=8.5}
df_yc <- yc$data |>
  mutate(
    maturity_label = paste0(maturity_years, "-year maturity"),
    maturity_label = factor(maturity_label,
                            levels = c("1-year maturity",
                                       "5-year maturity",
                                       "10-year maturity"))
  )

p_yc <- ggplot(df_yc, aes(x = date, y = yield_pct,
                          colour = maturity_label,
                          tooltip = maturity_label,
                          data_id = maturity_label)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line_interactive(linewidth = 0.3, show.legend = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  # Adaptive x-axis ticks: pretty_breaks gives ~6 ticks regardless of
  # the data range, and label_date_short formats them at the right
  # granularity (year for multi-year ranges; YYYY-MM for the zoom panel
  # which spans only the last 12 months). Avoids hardcoded date breaks.
  scale_x_date(breaks = scales::breaks_pretty(n = 6),
               labels = scales::label_date_short(format = c("%Y", "%m", "%d"))) +
  scale_colour_manual(values = mo_palette$categorical[1:3]) +
  ggforce::facet_zoom(xy = between(date, today_d - 365, today_d),
                      horizontal = FALSE, zoom.size = 0.5) +
  labs(x = NULL, y = "Yield",
       title = "Euro Area government bond yields",
       subtitle = "Daily spot rates, triple-A issuers only") +
  theme_mo()

girafe(ggobj = p_yc, width_svg = 8.5, height_svg = 4.5)
```

Lending money for longer is normally associated with higher yields; when
short-term yields overtake long-term yields, the yield curve has inverted,
which has historically been predictive of recessions.

# Government bond yields by country

```{r country-yields-prep, include=FALSE}
country_data <- eus$data |>
  filter(!geo %in% c("EA", "EU27_2020"))

# Latest observation per country, then rank for max/min prose. We must
# drop NAs BEFORE slice_min/max — those functions return zero-row
# results if the value column is all-NA in a group, which then crashes
# the inline R expressions further down with NULL$country_name.
cur_latest <- country_data |>
  filter(!is.na(yield_pct)) |>
  arrange(desc(date)) |>
  group_by(country_name) |>
  slice(1) |>
  ungroup()

cur_max_yld <- cur_latest |> slice_max(yield_pct, n = 1, with_ties = FALSE)
cur_min_yld <- cur_latest |> slice_min(yield_pct, n = 1, with_ties = FALSE)

# Euro Area average (last 12 months) and non-euro EU-member average.
ea_recent <- eus$data |>
  filter(geo == "EA", date >= today_d - 365) |>
  summarise(mean = round(mean(yield_pct, na.rm = TRUE), 2))

non_euro_geos <- c("CZ", "PL", "HU", "RO", "BG", "SE", "DK")
non_ea_recent <- country_data |>
  filter(geo %in% non_euro_geos, date >= today_d - 365) |>
  summarise(mean = round(mean(yield_pct, na.rm = TRUE), 2))
```

```{r country-yields-plot, echo=FALSE, fig.height=7, fig.width=9}
ea_series <- eus$data |> filter(geo == "EA") |> select(-geo, -country_name)

p_country <- ggplot(country_data,
                    aes(x = date, y = yield_pct)) +
  geom_line(data = ea_series, aes(x = date, y = yield_pct),
            colour = mo_palette$categorical[5], linetype = 2) +
  geom_line_interactive(colour = "grey55", linewidth = 0.3) +
  geom_point_interactive(
    aes(tooltip = paste0(country_name, ": ", round(yield_pct, 2), "% (",
                         format(date, "%Y-%m"), ")"),
        data_id  = paste0(geo, format(date, "-%Y-%m"))),
    size = 1.5, colour = "grey35"
  ) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  facet_wrap(~ country_name) +
  labs(x = NULL, y = "Yield",
       title = "Government bond yields in the EU",
       subtitle = "10-year maturity, monthly data",
       caption = "Dashed line: Euro Area average") +
  theme_mo() +
  theme(axis.text.x = element_text(size = 6),
        strip.text  = element_text(hjust = 1))

girafe(ggobj = p_country, width_svg = 9, height_svg = 7)
```

As most EU borrowing happens at the national level, yields vary
systematically by country as well as over time. At the moment the highest
yields are on bonds issued by **`r cur_max_yld$country_name`** at
`r mono(paste0(round(cur_max_yld$yield_pct, 2), "%"))`; the lowest are on
**`r cur_min_yld$country_name`** at
`r mono(paste0(round(cur_min_yld$yield_pct, 2), "%"))`.
Over the last twelve months, the Euro Area aggregate yield averaged
`r mono(paste0(ea_recent$mean, "%"))`, while EU members outside the euro
averaged `r mono(paste0(non_ea_recent$mean, "%"))`. The simplistic comparison is confounded by
uneven group composition (macroeconomic stability, debt loads), but
comparing matched pairs across the divide — e.g. Czechia vs. Slovakia, or
Finland vs. Sweden — can be instructive.

# Euro Area key interest rate and inflation

The ECB's primary mandate is price stability, with a long-term inflation
target of 2% year-on-year.

```{r rates-and-inflation, echo=FALSE, fig.height=4.5, fig.width=8.5}
# Join the two series on a daily-frequency axis so geom_step produces a
# continuous line. The previous version used vec_fill_missing on per-series
# data frames; if either source started later than 1998 the fill left a
# leading NA gap that filter(!is.na) then dropped, creating a visible
# discontinuity where the two series didn't share a start date. Now we
# join both series onto the SAME daily date scaffold and forward-fill in
# one pass so each series begins at its own first observation and runs
# unbroken to today.
rates_wide <- tibble::tibble(date = seq.Date(min(mrr$data$date, hic$data$date),
                                             today_d, by = "day")) |>
  left_join(mrr$data |> rename(mrr_pct = rate_pct), by = "date") |>
  left_join(hic$data |> rename(hicp_pct = yoy_pct), by = "date") |>
  arrange(date) |>
  mutate(`ECB main refinancing rate`    = vctrs::vec_fill_missing(mrr_pct, direction = "down"),
         `HICP year-on-year inflation`  = vctrs::vec_fill_missing(hicp_pct, direction = "down")) |>
  select(date, `ECB main refinancing rate`, `HICP year-on-year inflation`)

rates_long <- rates_wide |>
  tidyr::pivot_longer(cols = -date, names_to = "series", values_to = "value") |>
  filter(!is.na(value))

# y range for the zoom panel — set explicitly so the inner panel rescales
# to the values that actually appear in the last 12 months instead of
# inheriting the full historical 0-5% range.
rates_recent <- rates_long |> filter(date >= today_d - 365)
zoom_ylim <- range(rates_recent$value, na.rm = TRUE) + c(-0.3, 0.3)

p_rates <- ggplot(rates_long,
                  aes(x = date, y = value,
                      colour = series,
                      tooltip = paste0(series, ": ", round(value, 2), "% (",
                                       format(date, "%Y-%m"), ")"),
                      data_id = series)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_hline(yintercept = 2, colour = "grey50", linetype = 2) +
  geom_step_interactive(linewidth = 0.35, show.legend = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(breaks = scales::breaks_pretty(n = 6),
               labels = scales::label_date_short(format = c("%Y", "%m", "%d"))) +
  scale_colour_manual(values = mo_palette$categorical[c(2, 6)]) +
  ggforce::facet_zoom(xlim = c(today_d - 365, today_d),
                      ylim = zoom_ylim,
                      horizontal = FALSE, zoom.size = 0.5) +
  labs(x = NULL, y = "Rate",
       title = "Key interest rate and inflation",
       subtitle = "ECB main refinancing operations rate and Euro Area HICP year-on-year change",
       caption = "Dashed grey line marks the ECB's 2% inflation target.") +
  theme_mo()

girafe(ggobj = p_rates, width_svg = 8.5, height_svg = 4.5)
```

In times of rising inflation, the ECB's main countervailing instrument is
to raise key interest rates.

# Cite

Ovádek, M. *Borrowing and Yields in the European Union*. Accessed `r format(today_d, "%d %B %Y")`. Available at <https://michalovadek.github.io/euverse/trackers/eu-finance.html>.
