---
title: "Borrowing and Yields in the European Union"
subtitle: "An automatically updated overview of the EU's finances"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(ggplot2)
  library(ggiraph)
  library(ggforce)
  library(dplyr)
  library(lubridate)
  library(stringr)
  library(here)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load all four snapshots once at the top of the page.
yc  <- read_with_freshness("ecb_yield_curve")
eus <- read_with_freshness("eurostat_teimf050")
mrr <- read_with_freshness("ecb_mrr")
hic <- read_with_freshness("ecb_hicp")

today_d <- Sys.Date()
```

The European Union is not only a single market but also an economic and
monetary union. Most members share the euro, which entails a supranational
monetary policy run by the European Central Bank (ECB) together with Euro
Area central banks. This page provides an automatically updated overview of
several indicators central to that policy, sourced from the
[ECB Data Portal API](https://data.ecb.europa.eu/help/api/) and Eurostat.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(yc$meta, eus$meta, mrr$meta, hic$meta))
```

# Euro Area government bond yields

Government borrowing is normally carried out through bond sales. Bonds entitle
the buyer to an interest rate (coupon) and the return of the principal at
maturity. Bonds trade on secondary markets and their yields fluctuate
accordingly.

```{r euro-area-yields, echo=FALSE, fig.height=4.5, fig.width=8.5}
df_yc <- yc$data |>
  mutate(
    maturity_label = paste0(maturity_years, "-year maturity"),
    maturity_label = factor(maturity_label,
                            levels = c("1-year maturity",
                                       "5-year maturity",
                                       "10-year maturity"))
  )

p_yc <- ggplot(df_yc, aes(x = date, y = yield_pct,
                          colour = maturity_label,
                          tooltip = maturity_label,
                          data_id = maturity_label)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_line_interactive(linewidth = 0.3, show.legend = FALSE) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  # Adaptive x-axis ticks via date_breaks_min3 (in R/theme.R): always
  # returns at least 3 ticks regardless of range. The MAIN panel sees
  # the full 2004-present range and gets year-level breaks; the ZOOM
  # panel sees only the last 12 months and falls back to 2-month
  # breaks so the reader gets ≥3 anchor labels (was just "2026").
  scale_x_date(breaks = date_breaks_min3(),
               labels = scales::label_date_short(format = c("%Y", "%m", "%d"))) +
  scale_colour_manual(values = mo_palette$categorical[1:3]) +
  ggforce::facet_zoom(xy = between(date, today_d - 365, today_d),
                      horizontal = FALSE, zoom.size = 0.5) +
  labs(x = NULL, y = "Yield",
       title = "Euro Area government bond yields",
       subtitle = "Daily spot rates, triple-A issuers only") +
  theme_mo()

girafe(ggobj = p_yc, width_svg = 8.5, height_svg = 4.5)
```

Lending money for longer is normally associated with higher yields; when
short-term yields overtake long-term yields, the yield curve has inverted,
which has historically been predictive of recessions.

# Government bond yields by country

```{r country-yields-prep, include=FALSE}
country_data <- eus$data |>
  filter(!geo %in% c("EA", "EU27_2020"))

# Latest observation per country, then rank for max/min prose. We must
# drop NAs BEFORE slice_min/max because those functions return zero-row
# results if the value column is all-NA in a group, which then crashes
# the inline R expressions further down with NULL$country_name.
cur_latest <- country_data |>
  filter(!is.na(yield_pct)) |>
  arrange(desc(date)) |>
  group_by(country_name) |>
  slice(1) |>
  ungroup()

cur_max_yld <- cur_latest |> slice_max(yield_pct, n = 1, with_ties = FALSE)
cur_min_yld <- cur_latest |> slice_min(yield_pct, n = 1, with_ties = FALSE)

# Euro Area average (last 12 months) and non-euro EU-member average.
ea_recent <- eus$data |>
  filter(geo == "EA", date >= today_d - 365) |>
  summarise(mean = round(mean(yield_pct, na.rm = TRUE), 2))

non_euro_geos <- c("CZ", "PL", "HU", "RO", "BG", "SE", "DK")
non_ea_recent <- country_data |>
  filter(geo %in% non_euro_geos, date >= today_d - 365) |>
  summarise(mean = round(mean(yield_pct, na.rm = TRUE), 2))
```

```{r country-yields-plot, echo=FALSE, fig.height=7, fig.width=9}
ea_series <- eus$data |> filter(geo == "EA") |> select(-geo, -country_name)

p_country <- ggplot(country_data,
                    aes(x = date, y = yield_pct, colour = geo)) +
  geom_line(data = ea_series, aes(x = date, y = yield_pct),
            colour = "grey40", linetype = 2, inherit.aes = FALSE) +
  geom_line_interactive(linewidth = 0.4) +
  geom_point_interactive(
    aes(tooltip = paste0(country_name, ": ", round(yield_pct, 2), "% (",
                         format(date, "%Y-%m"), ")"),
        data_id  = paste0(geo, format(date, "-%Y-%m"))),
    size = 1.4
  ) +
  scale_x_date(breaks = date_breaks_min3(),
               labels = scales::label_date_short(format = c("%Y", "%m", "%d"))) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_colour_manual(values = mo_palette$ms_flags, guide = "none") +
  facet_wrap(~ country_name) +
  labs(x = NULL, y = "Yield",
       title = "Government bond yields in the EU",
       subtitle = "10-year maturity, monthly data",
       caption = "Dashed grey line: Euro Area average") +
  theme_mo() +
  theme(axis.text.x = element_text(size = 6),
        strip.text  = element_text(hjust = 1))

girafe(ggobj = p_country, width_svg = 9, height_svg = 7)
```

As most EU borrowing happens at the national level, yields vary
systematically by country as well as over time. At the moment the highest
yields are on bonds issued by **`r cur_max_yld$country_name`** at
`r mono(paste0(round(cur_max_yld$yield_pct, 2), "%"))`; the lowest are on
**`r cur_min_yld$country_name`** at
`r mono(paste0(round(cur_min_yld$yield_pct, 2), "%"))`.
Over the last twelve months, the Euro Area aggregate yield averaged
`r mono(paste0(ea_recent$mean, "%"))`, while EU members outside the euro
averaged `r mono(paste0(non_ea_recent$mean, "%"))`. The simplistic comparison is confounded by
uneven group composition (macroeconomic stability, debt loads), but
comparing matched pairs across the divide (for example Czechia vs.
Slovakia, or Finland vs. Sweden) can be instructive.

# Euro Area key interest rate and inflation

The ECB's primary mandate is price stability, with a long-term inflation
target of 2% year-on-year.

```{r rates-and-inflation, echo=FALSE, fig.height=4.5, fig.width=8.5}
# Combine the two indicators in their NATIVE frequency (MRR is event-
# driven, HICP monthly) and let geom_step draw the trajectory. The
# previous version daily-filled the data and used geom_line_interactive,
# which made ggiraph render ~700 separate polyline segments per series
# (instead of one continuous line per series), producing the visual
# fragmentation the reader sees. Event data + plain geom_step gives
# one clean step line per series.
rates_long <- bind_rows(
  mrr$data |> transmute(date = as.Date(date),
                        series = "ECB main refinancing rate",
                        value = rate_pct),
  hic$data |> transmute(date = as.Date(date),
                        series = "HICP year-on-year inflation",
                        value = yoy_pct)
) |>
  filter(!is.na(value)) |>
  arrange(series, date)

# Extend each series to today so the step line carries the last value
# to the right edge of the plot. Without this the step terminates at
# the last observation date with a gap to the panel border.
rates_long <- rates_long |>
  group_by(series) |>
  group_modify(~ if (max(.x$date) < today_d) {
    bind_rows(.x, tibble::tibble(date = today_d, value = tail(.x$value, 1)))
  } else .x) |>
  ungroup()

# y range for the zoom panel, set explicitly so the inner panel rescales
# to the values that actually appear in the last 12 months instead of
# inheriting the full historical 0-5% range.
rates_recent <- rates_long |> filter(date >= today_d - 365)
zoom_ylim <- range(rates_recent$value, na.rm = TRUE) + c(-0.3, 0.3)

p_rates <- ggplot(rates_long,
                  aes(x = date, y = value, colour = series)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_hline(yintercept = 2, colour = "grey50", linetype = 2) +
  # geom_step (non-interactive) gives one continuous polyline per
  # series. Interactivity lives on geom_point_interactive below,
  # which only renders at actual observations so tooltip targets
  # match the data the reader is looking at.
  geom_step(linewidth = 0.45, show.legend = FALSE) +
  geom_point_interactive(
    aes(tooltip = paste0(series, ": ", round(value, 2), "% (",
                         format(date, "%Y-%m"), ")"),
        data_id = paste0(series, format(date, "%Y-%m-%d"))),
    size = 0.9, alpha = 0.7, show.legend = FALSE
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(breaks = date_breaks_min3(),
               labels = scales::label_date_short(format = c("%Y", "%m", "%d"))) +
  scale_colour_manual(values = mo_palette$categorical[c(2, 6)]) +
  ggforce::facet_zoom(xlim = c(today_d - 365, today_d),
                      ylim = zoom_ylim,
                      horizontal = FALSE, zoom.size = 0.5) +
  labs(x = NULL, y = "Rate",
       title = "Key interest rate and inflation",
       subtitle = "ECB main refinancing operations rate and Euro Area HICP year-on-year change",
       caption = "Dashed grey line marks the ECB's 2% inflation target.") +
  theme_mo()

girafe(ggobj = p_rates, width_svg = 8.5, height_svg = 4.5)
```

In times of rising inflation, the ECB's main countervailing instrument is
to raise key interest rates.

# Cite

Ovádek, M. *Borrowing and Yields in the European Union*. Accessed `r format(today_d, "%d %B %Y")`. Available at <https://michalovadek.github.io/euverse/trackers/eu-finance.html>.
