---
title: "Legislative Output of the European Union"
subtitle: "An automatically updated overview of the EU's legislative activity"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(forcats)
  library(ggplot2)
  library(ggiraph)
  library(here)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
source(here::here("R", "celex.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load 4 nightly snapshots + 2 derived data-final masters.
acts_meta       <- read_with_freshness("eurlex_acts")
force_meta      <- read_with_freshness("eurlex_acts_force")
proposals_meta  <- read_with_freshness("eurlex_proposals")
lbs_meta        <- read_with_freshness("eurlex_lbs")
ms              <- arrow::read_parquet(here::here("data-final", "eu_member_states.parquet"))
# proposals_master: EUPROPS-trusted historical proposals (1958-2022) +
# Eur-Lex tail (post-2022). Carries proposal_type, is_amending, adopted
# date, withdrawn status — all of which the raw Eur-Lex proposals
# snapshot lacks. See R/build_proposals_master.R for the merge logic.
props_master  <- arrow::read_parquet(here::here("data-final", "proposals_master.parquet"))
prop_acts     <- arrow::read_parquet(here::here("data-final", "proposal_acts.parquet"))

acts       <- acts_meta$data
acts_force <- force_meta$data
proposals  <- proposals_meta$data
lbs        <- lbs_meta$data

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))

# Year x n_ms panel from shared master.
ms_count <- tibble(year = 1952:year_now) |>
  tidyr::crossing(ms) |>
  filter(year >= accession_year,
         is.na(exit_year) | year < exit_year) |>
  count(year, name = "n_ms")

# 4 main types.
acts_main <- acts |>
  filter(type %in% c("Regulation", "Directive", "Decision", "Recommendation"))

# Counts
n_regs <- sum(acts$type == "Regulation")
n_dirs <- sum(acts$type == "Directive")
n_decs <- sum(acts$type == "Decision")
n_recs <- sum(acts$type == "Recommendation")
```

The European Union is often criticised for having a heavy regulatory touch.
What is the reality? This page provides an automatically updated overview
of the EU's legislative output and efficiency, sourced from
[Eur-Lex](https://eur-lex.europa.eu/) via the
[eurlex](https://github.com/michalovadek/eurlex) R package.^[Any omissions
or mistakes in Eur-Lex carry through to the output shown here.]

As of `r format(today_d, "%d %B %Y")`, the EU and its predecessors have
produced `r mono(fmt_n(n_regs))` regulations,
`r mono(fmt_n(n_dirs))` directives,
`r mono(fmt_n(n_decs))` decisions, and
`r mono(fmt_n(n_recs))` recommendations.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(acts_meta$meta, force_meta$meta, proposals_meta$meta, lbs_meta$meta))
```

The following table shows the most recent legislation:

```{r recent-acts, echo=FALSE}
acts |>
  filter(!is.na(recent_title)) |>
  arrange(desc(as.Date(date)), desc(celex)) |>
  transmute(
    Date  = as.Date(date),
    Type  = type,
    CELEX = celex,
    Title = recent_title
  ) |>
  dt_mo()
```

# Number of acts over time

We start by looking at how the number of the four main legal acts changed
over the lifespan of the EU. The number adopted yearly is a parsimonious
proxy for the EU's regulatory tendencies.

```{r n-acts-time, echo=FALSE, fig.height=5, fig.width=9}
n_relevant <- acts_main |>
  count(year, type)

p_n_acts <- ggplot(n_relevant,
                   aes(x = year, y = n, fill = type,
                       tooltip = paste0(type, "s = ", n, " (", year, ")"),
                       data_id = paste0(type, year))) +
  geom_col_interactive(show.legend = FALSE) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  facet_wrap(~ type, scales = "free_y") +
  labs(x = NULL, y = "Number of acts",
       title = "Number of legal acts produced by the EU",
       subtitle = "Aggregated by year and type, by date of publication") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_n_acts, width_svg = 9, height_svg = 5)
```

# Proportion of act types

Some types of acts, notably regulations, have been historically more common
than others. We can visualise the evolution of proportions directly.

```{r prop-time, echo=FALSE, fig.height=4, fig.width=9}
proptime <- expand.grid(year = min(n_relevant$year):max(n_relevant$year),
                        type = unique(n_relevant$type)) |>
  left_join(n_relevant, by = c("year", "type")) |>
  mutate(n = ifelse(is.na(n), 0L, n)) |>
  group_by(year) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

p_prop <- ggplot(proptime, aes(x = year, y = proportion, fill = type)) +
  geom_area() +
  geom_vline(xintercept = c(1960, 1980, 2000, 2020), lty = 2, colour = "grey70") +
  geom_hline(yintercept = 0.50, colour = "grey50", lty = 3) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = NULL, y = NULL,
       title = "Relative prevalence of legal acts",
       subtitle = "Proportion of each type of legal act in a given year",
       fill = NULL) +
  theme_mo()

p_prop
```

Although regulations used to dominate the EU's legislative output,
decisions — typically associated with administrative actions — are nowadays
almost equally prevalent.

# Year-on-year comparison

We can look at the extent to which monthly adoption rates this year differ
from last year.

```{r yoy, echo=FALSE, fig.height=4, fig.width=9}
# One row per (act_type, calendar month) for current and previous year.
acts_monthly <- acts_main |>
  mutate(month_year = as.Date(paste0(str_sub(date, 1, 7), "-01"))) |>
  filter(year %in% c(year_now, year_now - 1)) |>
  count(month_year, type, name = "n")

# Months of the CURRENT year that have at least one observation. The
# chart shows one bar per (month, type) for those months only, so it
# grows naturally as new monthly data lands.
current_months <- acts_monthly |>
  filter(format(month_year, "%Y") == as.character(year_now)) |>
  distinct(month_year) |>
  arrange(month_year) |>
  mutate(month = format(month_year, "%m")) |>
  pull(month)

# YoY diff per (current-year month, type) = this year's count minus
# last year's count for the same calendar month.
yoy_diff <- expand.grid(month = current_months,
                        type  = unique(acts_main$type)) |>
  rowwise() |>
  mutate(n_now  = sum(acts_monthly$n[format(acts_monthly$month_year, "%m") == month &
                                     format(acts_monthly$month_year, "%Y") == as.character(year_now) &
                                     acts_monthly$type == type]),
         n_last = sum(acts_monthly$n[format(acts_monthly$month_year, "%m") == month &
                                     format(acts_monthly$month_year, "%Y") == as.character(year_now - 1) &
                                     acts_monthly$type == type]),
         yoy_diff = n_now - n_last) |>
  ungroup()

p_yoy <- ggplot(yoy_diff,
                aes(x = month, y = yoy_diff, fill = type,
                    tooltip = paste0(type, ": ", n_now, " in ", year_now,
                                     " vs ", n_last, " in ", year_now - 1,
                                     " (Δ ", yoy_diff, ")"),
                    data_id = paste0(type, month))) +
  geom_col_interactive(show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "grey83") +
  facet_wrap(~ type, scales = "free_y") +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  labs(x = NULL, y = "Year-on-year difference",
       title = paste0("Monthly year-on-year comparison: ", year_now, " vs ", year_now - 1),
       subtitle = paste0("Acts adopted in each month of ", year_now,
                         " minus the same month in ", year_now - 1)) +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_yoy, width_svg = 9, height_svg = 4)
```

# Acts in force

Not all legal acts remain in force indefinitely. According to Eur-Lex, the
EU has `r mono(fmt_n(sum(acts_force$force == 'true')))` acts
in force at the moment.

EU law tends to be cumulative: most acts adopted in any given year amend
or implement an earlier one. Data on this from EUPROPS (a curated
proposals dataset; see [Cite](#cite)) shows that roughly
`r mono(paste0(round(100 * mean(props_master$is_amending[!is.na(props_master$is_amending)]), 0), "%"))`
of Commission proposals on record are amending proposals.

```{r in-force, echo=FALSE, fig.height=4, fig.width=9}
in_force_clean <- acts_force |>
  filter(force == "true") |>
  mutate(days_in_force = as.integer(as.Date(today_d) - as.Date(date_force)),
         type = celex_act_type(celex)) |>
  filter(type != "Other")

p_inforce <- ggplot(in_force_clean,
                    aes(x = days_in_force, fill = type)) +
  geom_histogram(bins = 100, show.legend = FALSE) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  facet_wrap(~ type, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "How old are currently applicable legal acts?",
       subtitle = "Histogram of days since currently-applicable acts entered into force") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

p_inforce
```

# Proposals

Behind every adopted act is at least one Commission proposal. The
[EUPROPS dataset](https://www.michalovadek.github.io/) (curated through
2022) corrects systematic errors in Eur-Lex's proposal metadata; we
combine it with the nightly Eur-Lex pull for the post-2022 tail. Total
proposals in scope: `r mono(fmt_n(nrow(props_master)))`, of which
`r mono(fmt_n(sum(!is.na(props_master$adopted))))` led to at least one
adopted act and
`r mono(fmt_n(sum(props_master$withdrawn %in% c('withdrawn', 'rejected', 'not adopted', 'replaced'))))`
were withdrawn, rejected, replaced, or never adopted.

```{r prop-by-type, echo=FALSE, fig.height=4.5, fig.width=9}
props_yearly <- props_master |>
  filter(!is.na(proposed),
         proposal_type %in% c("Regulation", "Directive", "Decision", "Recommendation")) |>
  mutate(year = as.integer(format(proposed, "%Y"))) |>
  filter(year >= 1958, year <= year_now) |>
  count(year, proposal_type)

p_props <- ggplot(props_yearly,
                  aes(x = year, y = n, fill = proposal_type,
                      tooltip = paste0(proposal_type, ": ", n, " proposals (", year, ")"),
                      data_id = paste0(proposal_type, year))) +
  geom_col_interactive(show.legend = FALSE) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  facet_wrap(~ proposal_type, scales = "free_y") +
  labs(x = NULL, y = NULL,
       title = "Number of Commission proposals over time",
       subtitle = "By year of proposal and type") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_props, width_svg = 9, height_svg = 4.5)
```

Decisions are by far the most numerous category, while directives are
the rarest — reflecting their heavier political weight (a directive
binds Member States as to result but lets them choose form and methods).

## Outcomes

Not every proposal makes it into the EU's legal corpus. Many are
withdrawn before adoption — sometimes because of Council deadlock,
sometimes because the Commission has decided to redraft.

```{r prop-outcomes, echo=FALSE, fig.height=4, fig.width=9}
outcomes <- props_master |>
  filter(source_data == "euprops",  # outcome status only reliable for EUPROPS
         !is.na(proposed),
         proposal_type %in% c("Regulation", "Directive", "Decision", "Recommendation")) |>
  mutate(year = as.integer(format(proposed, "%Y")),
         outcome = case_when(
           !is.na(adopted)                 ~ "Adopted",
           withdrawn == "withdrawn"        ~ "Withdrawn",
           withdrawn == "rejected"         ~ "Rejected",
           withdrawn == "replaced"         ~ "Replaced",
           withdrawn == "not adopted"      ~ "Not adopted",
           withdrawn == "repealed"         ~ "Repealed",
           TRUE                            ~ "Other / pending"
         )) |>
  filter(year >= 1980, year <= year_now) |>
  count(year, outcome) |>
  group_by(year) |>
  mutate(share = n / sum(n)) |>
  ungroup() |>
  mutate(outcome = factor(outcome,
                          levels = c("Adopted", "Withdrawn", "Replaced",
                                     "Rejected", "Not adopted", "Repealed",
                                     "Other / pending")))

p_outcomes <- ggplot(outcomes, aes(x = year, y = share, fill = outcome,
                                   tooltip = paste0(outcome, ": ", n, " (", year, ")"),
                                   data_id = paste0(outcome, year))) +
  ggiraph::geom_col_interactive(show.legend = TRUE) +
  scale_fill_manual(values = c(Adopted        = mo_palette$categorical[3],
                               Withdrawn      = mo_palette$categorical[6],
                               Replaced       = mo_palette$categorical[5],
                               Rejected       = mo_palette$categorical[1],
                               `Not adopted`  = mo_palette$categorical[7],
                               Repealed       = mo_palette$categorical[2],
                               `Other / pending` = mo_palette$categorical[8])) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Outcome of Commission proposals over time",
       subtitle = "Share of proposals by status, EUPROPS-curated data (1980-2022)") +
  theme_mo()

girafe(ggobj = p_outcomes, width_svg = 9, height_svg = 4)
```

The vast majority of proposals end up adopted; withdrawal/rejection
shares hover around the low single digits in most years.

## Amending vs new proposals

EUPROPS distinguishes "amending" proposals (which modify an existing
act) from genuinely new ones. The amending share is a rough indicator
of how much of the legislative effort goes into evolving the existing
corpus versus opening new fronts.

```{r amending-share, echo=FALSE, fig.height=4, fig.width=9}
am_share <- props_master |>
  filter(source_data == "euprops", !is.na(is_amending), !is.na(proposed),
         proposal_type %in% c("Regulation", "Directive", "Decision")) |>
  mutate(year = as.integer(format(proposed, "%Y")),
         category = ifelse(is_amending, "Amending", "New")) |>
  filter(year >= 1980, year <= year_now) |>
  count(year, proposal_type, category) |>
  group_by(year, proposal_type) |>
  mutate(share = n / sum(n)) |>
  ungroup()

p_am <- ggplot(am_share, aes(x = year, y = share, fill = category,
                             tooltip = paste0(category, ": ", n, " (", year, ", ", proposal_type, ")"),
                             data_id = paste0(category, year, proposal_type))) +
  ggiraph::geom_col_interactive(show.legend = TRUE) +
  scale_fill_manual(values = c(Amending = mo_palette$categorical[6],
                               New      = mo_palette$categorical[5])) +
  scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  facet_wrap(~ proposal_type, ncol = 1, scales = "fixed") +
  labs(x = NULL, y = NULL, fill = NULL,
       title = "Amending vs new Commission proposals",
       subtitle = "Share by year and proposal type, EUPROPS-curated data (1980-2022)") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_am, width_svg = 9, height_svg = 6)
```

# Legislative efficiency

The image of the EU as an unwieldy bureaucracy is widespread. Let's use
EUPROPS-corrected dates (more reliable than raw Eur-Lex for older
proposals) to find out how long it normally takes to pass legislation.

```{r efficiency, echo=FALSE}
# Days from proposal to adoption, drawn from EUPROPS where both dates
# are present. EUPROPS dates are the curated source of truth, so we no
# longer need the acts + proposals join + min-date scramble the old
# Eur-Lex-only pipeline used. Restrict to the 3 main act types since
# recommendations are too few for stable inference.
days_master <- props_master |>
  filter(source_data == "euprops",
         !is.na(proposed), !is.na(adopted),
         proposal_type %in% c("Regulation", "Directive", "Decision")) |>
  mutate(days         = as.integer(adopted - proposed),
         year_adopted = as.integer(format(adopted, "%Y"))) |>
  filter(days > 0, year_adopted > 1985)

# Outliers = > 2.9× the within-type mean. Used for both the scatter plot
# below AND excluded from the regression (so the model isn't driven by
# 20-year-stuck files).
outliers <- days_master |>
  group_by(proposal_type) |>
  mutate(mean = mean(days), coef = days / mean) |>
  filter(coef > 2.9) |>
  ungroup()
```

```{r outlier-scatter, echo=FALSE, fig.height=5, fig.width=9}
outliers_top <- outliers |>
  arrange(desc(days)) |>
  slice(1:50) |>
  mutate(title_clean = str_squish(str_remove(title, "\\(?Text with EEA relevance\\)?")),
         tooltip = ifelse(is.na(title_clean) | title_clean == "",
                          paste0(euprop_id, ": ", days, " days"),
                          paste0(title_clean, " (", days, " days, ",
                                 ifelse(is_amending, "amending", "new"), ")")),
         data_id = euprop_id,
         category = ifelse(is_amending, "Amending", "New"))

p_outliers <- ggplot(outliers_top,
                     aes(x = adopted, y = days,
                         fill = proposal_type, colour = proposal_type,
                         shape = category,
                         tooltip = tooltip, data_id = data_id)) +
  geom_point_interactive(alpha = 0.7, size = 2.7, show.legend = TRUE) +
  geom_hline(yintercept = c(3650, 7300), lty = 2, colour = "grey75") +
  annotate("text", y = c(3650 + 110, 7300 + 110), x = as.Date("2012-01-01"),
           label = c("10 years", "20 years"),
           colour = "grey50", fontface = "italic") +
  scale_fill_manual(values = mo_palette$categorical[1:3]) +
  scale_colour_manual(values = mo_palette$categorical[1:3]) +
  scale_shape_manual(values = c(Amending = 24, New = 21)) +
  labs(x = "Date of adoption", y = "Days to adoption",
       title = "The most slowly adopted acts in EU history",
       subtitle = "Days between Commission proposal and adoption (EUPROPS-curated dates)",
       caption = "Hover over points for title; triangles = amending proposals, circles = new",
       fill = NULL, colour = NULL, shape = NULL) +
  theme_mo()

girafe(ggobj = p_outliers, width_svg = 9, height_svg = 5)
```

EUPROPS lets us separate amending proposals (modifying an existing act)
from new ones — a control that wasn't possible with raw Eur-Lex data.

```{r efficiency-regression, echo=FALSE}
days_ms <- days_master |>
  filter(!euprop_id %in% outliers$euprop_id) |>
  left_join(ms_count, by = c("year_adopted" = "year")) |>
  mutate(proposal_type = relevel(as.factor(proposal_type), ref = "Decision"),
         amending      = ifelse(is_amending, "amending", "new"))

m_ms       <- lm(days ~ n_ms, data = days_ms)
m_with_typ <- lm(days ~ n_ms + proposal_type, data = days_ms)
m_full     <- lm(days ~ n_ms + proposal_type + amending, data = days_ms)

suppressPackageStartupMessages(library(modelsummary))
modelsummary::modelsummary(
  list("Baseline" = m_ms, "+ Type" = m_with_typ, "+ Amending flag" = m_full),
  statistic  = "conf.int",
  conf_level = 0.95,
  stars      = TRUE,
  title      = "Linear model of legislative efficiency (days to adoption)"
)
```

Adding the amending flag to the type controls shows that amending
proposals do not move significantly faster than new ones once type is
held constant — contrary to a common intuition that "just tweaking an
existing act" should be quicker. The Member-State coefficient is stable
across specifications: each additional Member State adds a few days to
average adoption time.

## Proposals that produced more than one act

A small but interesting share of proposals are split during the
legislative procedure and yield multiple adopted acts. EUPROPS records
this explicitly (`adoptedlaws` column, ";"-separated). We expose it as
a long table in `data-final/proposal_acts.parquet` so each linkage is a
row.

```{r multi-act, echo=FALSE}
multi <- prop_acts |>
  count(proposal_celex, euprop_id, name = "n_acts") |>
  filter(n_acts > 1) |>
  arrange(desc(n_acts)) |>
  left_join(props_master |> select(euprop_id, title, proposed, proposal_type),
            by = "euprop_id") |>
  transmute(
    `EUPROPID`     = euprop_id,
    `Proposal CELEX` = proposal_celex,
    `Proposed`     = proposed,
    `Type`         = proposal_type,
    `# acts`       = n_acts,
    `Title`        = str_squish(str_remove(title, "\\(?Text with EEA relevance\\)?"))
  )

dt_mo(multi)
```

# Legal bases

Each EU legal act has a _legal basis_ in the Treaties or in pre-existing
EU law. Looking at the most-invoked legal bases gives insight into EU
legislative activity.

```{r lbs-data, echo=FALSE}
lbs_clean <- lbs |>
  filter(!is.na(celex), !is.na(lbcelex)) |>
  mutate(
    delegated = ifelse(str_detect(lbcelex, "^1.*([:digit:]|[:punct:])$"), "primary", "delegated"),
    act_year  = as.integer(str_sub(celex, 2, 5)),
    lb_year   = as.integer(str_sub(lbcelex, 2, 5))
  )

most_common_lbs <- lbs_clean |>
  distinct(celex, lbcelex) |>
  count(lbcelex, sort = TRUE) |>
  slice(1:10) |>
  mutate(title = purrr::map_chr(
    str_c("http://publications.europa.eu/resource/celex/", lbcelex),
    purrr::possibly(eurlex::elx_fetch_data, otherwise = NA_character_),
    "title"
  )) |>
  mutate(title = str_squish(str_remove(title, "\\(?Text with EEA relevance\\)?")))

top_lb_name <- most_common_lbs$title[1]
top_lb_n    <- most_common_lbs$n[1]
```

The most-invoked legal basis historically is *`r if (is.na(top_lb_name)) most_common_lbs$lbcelex[1] else top_lb_name`*, used as a legal basis on `r mono(fmt_n(top_lb_n))` occasions.

Nonetheless, when we think of legal bases we usually imagine a Treaty
competence. Filtering to legal bases in the Treaty on the Functioning of
the European Union (TFEU) and Treaty on European Union (TEU), the
following table lists the most-invoked Treaty competences since the
entry into force of the Treaty of Lisbon.

```{r lbs-table, echo=FALSE}
# Lisbon-treaty CELEX format: 12008E… (TFEU), 12008M… (TEU). Mirror the
# table layout from the standalone tracker: human-readable "Article XX
# TFEU/TEU" instead of opaque CELEX codes.
lisbon_lbs <- lbs_clean |>
  distinct(celex, lbcelex) |>
  count(lbcelex, sort = TRUE) |>
  filter(str_detect(lbcelex, "^12008[EM][0-9]")) |>
  mutate(art_num    = as.integer(str_extract(lbcelex, "(?<=E|M)[0-9]+")),
         art_treaty = ifelse(str_detect(lbcelex, "^12008E"), "TFEU", "TEU"),
         `Treaty article` = paste0("Article ", art_num, " ", art_treaty)) |>
  transmute(`Treaty article`, `CELEX` = lbcelex, `Times invoked` = n)

dt_mo(lisbon_lbs)
```

```{r lbs-deleg-time, echo=FALSE, fig.height=4, fig.width=9}
deleg_prop_time <- lbs_clean |>
  distinct(celex, .keep_all = TRUE) |>
  count(act_year, delegated) |>
  complete(act_year, delegated, fill = list(n = 0L)) |>
  group_by(act_year) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

p_deleg <- ggplot(deleg_prop_time, aes(x = act_year, y = prop, fill = delegated)) +
  geom_area() +
  scale_fill_manual(values = mo_palette$categorical[c(1, 5)]) +
  geom_vline(xintercept = c(1960, 1980, 2000, 2020), lty = 2, colour = "grey70") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = NULL, y = NULL,
       title = "Relative prevalence of delegated and primary acts",
       subtitle = "Proportion of delegated vs primary acts in a given year",
       fill = NULL) +
  theme_mo()

p_deleg
```

Most legal acts adopted by the EU are delegated and implementing acts —
based on another piece of legislation rather than directly on a Treaty.

# Cite

Ovádek, M. *Legislative Output of the European Union*. Accessed
`r format(Sys.Date(), "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-law.html>.
