---
title: "Legislative Output of the European Union"
subtitle: "An automatically updated overview of the EU's legislative activity"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(forcats)
  library(ggplot2)
  library(ggiraph)
  library(here)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
source(here::here("R", "celex.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load 4 snapshots + shared eu_member_states.
acts_meta       <- read_with_freshness("eurlex_acts")
force_meta      <- read_with_freshness("eurlex_acts_force")
proposals_meta  <- read_with_freshness("eurlex_proposals")
lbs_meta        <- read_with_freshness("eurlex_lbs")
ms              <- arrow::read_parquet(here::here("data-final", "eu_member_states.parquet"))

acts      <- acts_meta$data
acts_force<- force_meta$data
proposals <- proposals_meta$data
lbs       <- lbs_meta$data

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))

# Year x n_ms panel from shared master.
ms_count <- tibble(year = 1952:year_now) |>
  tidyr::crossing(ms) |>
  filter(year >= accession_year,
         is.na(exit_year) | year < exit_year) |>
  count(year, name = "n_ms")

# 4 main types.
acts_main <- acts |>
  filter(type %in% c("Regulation", "Directive", "Decision", "Recommendation"))

# Counts
n_regs <- sum(acts$type == "Regulation")
n_dirs <- sum(acts$type == "Directive")
n_decs <- sum(acts$type == "Decision")
n_recs <- sum(acts$type == "Recommendation")
```

The European Union is often criticised for having a heavy regulatory touch.
What is the reality? This page provides an automatically updated overview
of the EU's legislative output and efficiency, sourced from
[Eur-Lex](https://eur-lex.europa.eu/) via the
[eurlex](https://github.com/michalovadek/eurlex) R package.^[Any omissions
or mistakes in Eur-Lex carry through to the output shown here.]

As of `r format(today_d, "%d %B %Y")`, the EU and its predecessors have
produced `r mono(fmt_n(n_regs))` regulations,
`r mono(fmt_n(n_dirs))` directives,
`r mono(fmt_n(n_decs))` decisions, and
`r mono(fmt_n(n_recs))` recommendations.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(acts_meta$meta, force_meta$meta, proposals_meta$meta, lbs_meta$meta))
```

The following table shows the most recent legislation:

```{r recent-acts, echo=FALSE}
acts |>
  filter(!is.na(recent_title)) |>
  arrange(desc(as.Date(date)), desc(celex)) |>
  transmute(
    Date  = as.Date(date),
    Type  = type,
    CELEX = celex,
    Title = recent_title
  ) |>
  dt_mo()
```

# Number of acts over time

We start by looking at how the number of the four main legal acts changed
over the lifespan of the EU. The number adopted yearly is a parsimonious
proxy for the EU's regulatory tendencies.

```{r n-acts-time, echo=FALSE, fig.height=5, fig.width=9}
n_relevant <- acts_main |>
  count(year, type)

p_n_acts <- ggplot(n_relevant,
                   aes(x = year, y = n, fill = type,
                       tooltip = paste0(type, "s = ", n, " (", year, ")"),
                       data_id = paste0(type, year))) +
  geom_col_interactive(show.legend = FALSE) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  facet_wrap(~ type, scales = "free_y") +
  labs(x = NULL, y = "Number of acts",
       title = "Number of legal acts produced by the EU",
       subtitle = "Aggregated by year and type, by date of publication") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_n_acts, width_svg = 9, height_svg = 5)
```

# Proportion of act types

Some types of acts, notably regulations, have been historically more common
than others. We can visualise the evolution of proportions directly.

```{r prop-time, echo=FALSE, fig.height=4, fig.width=9}
proptime <- expand.grid(year = min(n_relevant$year):max(n_relevant$year),
                        type = unique(n_relevant$type)) |>
  left_join(n_relevant, by = c("year", "type")) |>
  mutate(n = ifelse(is.na(n), 0L, n)) |>
  group_by(year) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()

p_prop <- ggplot(proptime, aes(x = year, y = proportion, fill = type)) +
  geom_area() +
  geom_vline(xintercept = c(1960, 1980, 2000, 2020), lty = 2, colour = "grey70") +
  geom_hline(yintercept = 0.50, colour = "grey50", lty = 3) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = NULL, y = NULL,
       title = "Relative prevalence of legal acts",
       subtitle = "Proportion of each type of legal act in a given year",
       fill = NULL) +
  theme_mo()

p_prop
```

Although regulations used to dominate the EU's legislative output,
decisions — typically associated with administrative actions — are nowadays
almost equally prevalent.

# Year-on-year comparison

We can look at the extent to which monthly adoption rates this year differ
from last year.

```{r yoy, echo=FALSE, fig.height=4, fig.width=9}
# One row per (act_type, calendar month) for current and previous year.
acts_monthly <- acts_main |>
  mutate(month_year = as.Date(paste0(str_sub(date, 1, 7), "-01"))) |>
  filter(year %in% c(year_now, year_now - 1)) |>
  count(month_year, type, name = "n")

# Months of the CURRENT year that have at least one observation. The
# chart shows one bar per (month, type) for those months only, so it
# grows naturally as new monthly data lands.
current_months <- acts_monthly |>
  filter(format(month_year, "%Y") == as.character(year_now)) |>
  distinct(month_year) |>
  arrange(month_year) |>
  mutate(month = format(month_year, "%m")) |>
  pull(month)

# YoY diff per (current-year month, type) = this year's count minus
# last year's count for the same calendar month.
yoy_diff <- expand.grid(month = current_months,
                        type  = unique(acts_main$type)) |>
  rowwise() |>
  mutate(n_now  = sum(acts_monthly$n[format(acts_monthly$month_year, "%m") == month &
                                     format(acts_monthly$month_year, "%Y") == as.character(year_now) &
                                     acts_monthly$type == type]),
         n_last = sum(acts_monthly$n[format(acts_monthly$month_year, "%m") == month &
                                     format(acts_monthly$month_year, "%Y") == as.character(year_now - 1) &
                                     acts_monthly$type == type]),
         yoy_diff = n_now - n_last) |>
  ungroup()

p_yoy <- ggplot(yoy_diff,
                aes(x = month, y = yoy_diff, fill = type,
                    tooltip = paste0(type, ": ", n_now, " in ", year_now,
                                     " vs ", n_last, " in ", year_now - 1,
                                     " (Δ ", yoy_diff, ")"),
                    data_id = paste0(type, month))) +
  geom_col_interactive(show.legend = FALSE) +
  geom_hline(yintercept = 0, colour = "grey83") +
  facet_wrap(~ type, scales = "free_y") +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  labs(x = NULL, y = "Year-on-year difference",
       title = paste0("Monthly year-on-year comparison: ", year_now, " vs ", year_now - 1),
       subtitle = paste0("Acts adopted in each month of ", year_now,
                         " minus the same month in ", year_now - 1)) +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

girafe(ggobj = p_yoy, width_svg = 9, height_svg = 4)
```

# Acts in force

Not all legal acts remain in force indefinitely. According to Eur-Lex, the
EU has `r mono(fmt_n(sum(acts_force$force == 'true')))` acts
in force at the moment.

```{r in-force, echo=FALSE, fig.height=4, fig.width=9}
in_force_clean <- acts_force |>
  filter(force == "true") |>
  mutate(days_in_force = as.integer(as.Date(today_d) - as.Date(date_force)),
         type = celex_act_type(celex)) |>
  filter(type != "Other")

p_inforce <- ggplot(in_force_clean,
                    aes(x = days_in_force, fill = type)) +
  geom_histogram(bins = 100, show.legend = FALSE) +
  scale_fill_manual(values = mo_palette$categorical[1:4]) +
  facet_wrap(~ type, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "How old are currently applicable legal acts?",
       subtitle = "Histogram of days since currently-applicable acts entered into force") +
  theme_mo() +
  theme(strip.text = element_text(hjust = 0, face = "bold"))

p_inforce
```

# Legislative efficiency

The image of the EU as an unwieldy bureaucracy is widespread. Let's use
Eur-Lex data to find out how long it normally takes to pass legislation.

```{r efficiency, echo=FALSE}
acts_days <- acts |>
  filter(!is.na(date), !is.na(proposal)) |>
  left_join(proposals, by = c("proposal" = "celex")) |>
  mutate(days = as.integer(as.Date(date) - as.Date(date_proposal))) |>
  filter(days > -1) |>
  group_by(celex) |>
  filter(days == max(days)) |>
  ungroup() |>
  distinct(celex, .keep_all = TRUE) |>
  arrange(date) |>
  mutate(year_adopted = as.integer(str_sub(date, 1, 4))) |>
  filter(year_adopted > 1985,
         type %in% c("Regulation", "Directive", "Decision"))

outliers <- acts_days |>
  group_by(type) |>
  mutate(mean = mean(days), coef = days / mean) |>
  filter(coef > 2.9) |>
  ungroup()
```

```{r outlier-scatter, echo=FALSE, fig.height=5, fig.width=9}
outliers_top <- outliers |>
  arrange(desc(days)) |>
  slice(1:50) |>
  select(-any_of("recent_title")) |>
  left_join(acts |> select(celex, recent_title), by = "celex") |>
  # The snapshot only carries titles for the top-50 newest acts. The
  # "most slowly adopted" outliers are mostly old, so recent_title is
  # NA for most of them and the tooltip falls back to a bare CELEX.
  # Fetch titles for any missing ones at render time (≤50 calls,
  # cached by Quarto's freeze).
  mutate(recent_title = ifelse(
    is.na(recent_title),
    purrr::map_chr(
      celex,
      ~ tryCatch(
          as.character(eurlex::elx_fetch_data(
            paste0("http://publications.europa.eu/resource/celex/", .x), "title"))[1],
          error = function(e) NA_character_)
    ),
    recent_title
  )) |>
  mutate(recent_title = str_squish(str_remove(recent_title, "\\(?Text with EEA relevance\\)?")),
         tooltip = ifelse(is.na(recent_title) | recent_title == "",
                          paste0(celex, ": ", days, " days"),
                          paste0(recent_title, " (", days, " days)")),
         data_id = celex)

p_outliers <- ggplot(outliers_top, aes(x = as.Date(date), y = days, fill = type, colour = type,
                                       tooltip = tooltip, data_id = data_id)) +
  geom_point_interactive(alpha = 0.6, size = 2.5, show.legend = TRUE) +
  geom_hline(yintercept = c(3650, 7300), lty = 2, colour = "grey75") +
  annotate("text", y = c(3650 + 110, 7300 + 110), x = as.Date("2012-01-01"),
           label = c("10 years", "20 years"),
           colour = "grey50", fontface = "italic") +
  scale_fill_manual(values = mo_palette$categorical[1:3]) +
  scale_colour_manual(values = mo_palette$categorical[1:3]) +
  labs(x = "Date of adoption", y = "Days to adoption",
       title = "The most slowly adopted acts in EU history",
       subtitle = "Number of days between initial Commission proposal and adoption",
       caption = "Hover over points for title",
       fill = NULL, colour = NULL) +
  theme_mo()

girafe(ggobj = p_outliers, width_svg = 9, height_svg = 5)
```

```{r efficiency-regression, echo=FALSE}
days_ms <- acts_days |>
  filter(!celex %in% outliers$celex) |>
  left_join(ms_count, by = c("year_adopted" = "year")) |>
  mutate(type = relevel(as.factor(type), ref = "Decision"))

m_ms      <- lm(days ~ n_ms, data = days_ms)
m_full    <- lm(days ~ n_ms + type, data = days_ms)

suppressPackageStartupMessages(library(modelsummary))
modelsummary::modelsummary(
  list("Baseline" = m_ms, "With type controls" = m_full),
  statistic  = "conf.int",
  conf_level = 0.95,
  stars      = TRUE,
  title      = "Linear model of legislative efficiency (days to adoption)"
)
```

Controlling for type, the model suggests that for every additional Member
State, the EU takes on average a few days longer to adopt a legal act.

# Legal bases

Each EU legal act has a _legal basis_ in the Treaties or in pre-existing
EU law. Looking at the most-invoked legal bases gives insight into EU
legislative activity.

```{r lbs-data, echo=FALSE}
lbs_clean <- lbs |>
  filter(!is.na(celex), !is.na(lbcelex)) |>
  mutate(
    delegated = ifelse(str_detect(lbcelex, "^1.*([:digit:]|[:punct:])$"), "primary", "delegated"),
    act_year  = as.integer(str_sub(celex, 2, 5)),
    lb_year   = as.integer(str_sub(lbcelex, 2, 5))
  )

most_common_lbs <- lbs_clean |>
  distinct(celex, lbcelex) |>
  count(lbcelex, sort = TRUE) |>
  slice(1:10) |>
  mutate(title = purrr::map_chr(
    str_c("http://publications.europa.eu/resource/celex/", lbcelex),
    purrr::possibly(eurlex::elx_fetch_data, otherwise = NA_character_),
    "title"
  )) |>
  mutate(title = str_squish(str_remove(title, "\\(?Text with EEA relevance\\)?")))

top_lb_name <- most_common_lbs$title[1]
top_lb_n    <- most_common_lbs$n[1]
```

The most-invoked legal basis historically is *`r if (is.na(top_lb_name)) most_common_lbs$lbcelex[1] else top_lb_name`*, used as a legal basis on `r mono(fmt_n(top_lb_n))` occasions.

Nonetheless, when we think of legal bases we usually imagine a Treaty
competence. Filtering to legal bases in the Treaty on the Functioning of
the European Union (TFEU) and Treaty on European Union (TEU), the
following table lists the most-invoked Treaty competences since the
entry into force of the Treaty of Lisbon.

```{r lbs-table, echo=FALSE}
# Lisbon-treaty CELEX format: 12008E… (TFEU), 12008M… (TEU). Mirror the
# table layout from the standalone tracker: human-readable "Article XX
# TFEU/TEU" instead of opaque CELEX codes.
lisbon_lbs <- lbs_clean |>
  distinct(celex, lbcelex) |>
  count(lbcelex, sort = TRUE) |>
  filter(str_detect(lbcelex, "^12008[EM][0-9]")) |>
  mutate(art_num    = as.integer(str_extract(lbcelex, "(?<=E|M)[0-9]+")),
         art_treaty = ifelse(str_detect(lbcelex, "^12008E"), "TFEU", "TEU"),
         `Treaty article` = paste0("Article ", art_num, " ", art_treaty)) |>
  transmute(`Treaty article`, `CELEX` = lbcelex, `Times invoked` = n)

dt_mo(lisbon_lbs)
```

```{r lbs-deleg-time, echo=FALSE, fig.height=4, fig.width=9}
deleg_prop_time <- lbs_clean |>
  distinct(celex, .keep_all = TRUE) |>
  count(act_year, delegated) |>
  complete(act_year, delegated, fill = list(n = 0L)) |>
  group_by(act_year) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

p_deleg <- ggplot(deleg_prop_time, aes(x = act_year, y = prop, fill = delegated)) +
  geom_area() +
  scale_fill_manual(values = mo_palette$categorical[c(1, 5)]) +
  geom_vline(xintercept = c(1960, 1980, 2000, 2020), lty = 2, colour = "grey70") +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = NULL, y = NULL,
       title = "Relative prevalence of delegated and primary acts",
       subtitle = "Proportion of delegated vs primary acts in a given year",
       fill = NULL) +
  theme_mo()

p_deleg
```

Most legal acts adopted by the EU are delegated and implementing acts —
based on another piece of legislation rather than directly on a Treaty.

# Cite

Ovádek, M. *Legislative Output of the European Union*. Accessed
`r format(Sys.Date(), "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-law.html>.
