---
title: "Politics of EU Member State Governments"
subtitle: "Ideological composition of the Council over time and the EU as a campaign issue"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggiraph)
  library(here)
  library(stringr)
  library(forcats)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Pre-built parquets from R/build_eu_politics.R - regenerated nightly
# from ParlGov view_cabinet + view_party joined to data-manual/eu-member-states
# and from the Manifesto Project MPDS dump.
ec <- arrow::read_parquet(
  here::here("data-final", "eu_council_aggregate.parquet")
)
eg <- arrow::read_parquet(
  here::here("data-final", "eu_government_composition.parquet")
)
em_eu <- arrow::read_parquet(
  here::here("data-final", "eu_manifesto_eu_stance_eu.parquet")
)
em_co <- arrow::read_parquet(
  here::here("data-final", "eu_manifesto_eu_stance_country.parquet")
)

ec_meta <- manual_meta(
  here::here("data-final", "eu_council_aggregate.parquet"),
  source_label = "ParlGov view_cabinet + view_party (built nightly)"
)
ec_meta$rows <- nrow(ec)
em_meta <- manual_meta(
  here::here("data-final", "eu_manifesto_eu_stance_eu.parquet"),
  source_label = "Manifesto Project MPDS 2024a (static drop-in)"
)
em_meta$rows <- nrow(em_co)

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))
```

The euverse's other trackers describe what EU governments _do_ — vetoes
they cast, laws they pass, cases they litigate. This page is about
_what governments are_: the partisan and ideological composition of
each Member State's national executive over time, aggregated up to the
Council of the European Union as a whole.

Two complementary sources sit underneath:

- **[ParlGov](https://www.parlgov.org/)** for the factual record of
  which parties were in cabinet in each Member State, when and with
  what share of seats. ParlGov also assigns each party a fixed
  left-right score and a pro/anti-EU score (`eu_anti_pro`, 0 = anti-EU,
  10 = pro-EU) drawn from expert surveys.
- **The [Manifesto Project (MPDS)](https://manifesto-project.wzb.eu/)**
  for what parties actually wrote about Europe in their election
  manifestos, election by election since 1945.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(ec_meta, em_meta))
```

# Ideological composition of the Council

The Council of the EU is composed of national governments; one seat
per Member State, regardless of population. The chart below summarises
that body's centre of gravity over time by:

1. Computing a **seat-weighted** ideological score for every cabinet
   (each cabinet party weighted by its parliamentary seats), separately
   for the left-right and pro/anti-EU dimensions.
2. **Equal-weighting** those cabinet scores across all Member States in
   each year, since the Council itself is one-MS-one-seat.

Caveats up front: this is "average national government", not a
Council-formation-specific view (we'd need minister-portfolio-party
data, which ParlGov doesn't provide), and ParlGov's `eu_anti_pro` field
is missing entirely for parties in **Luxembourg, Cyprus and Malta**,
so those three countries drop out of the EU-stance average (the
left-right average covers all 27/28).

```{r council-aggregate, echo=FALSE, fig.height=5, fig.width=9}
ec_long <- ec |>
  select(year, n_ms_total,
         `Left-right`  = council_left_right,
         `Pro/anti-EU` = council_eu_anti_pro) |>
  tidyr::pivot_longer(c(`Left-right`, `Pro/anti-EU`),
                      names_to = "dimension", values_to = "score") |>
  mutate(tooltip = paste0(year, " — ", dimension, ": ",
                          round(score, 2), "/10",
                          " (", n_ms_total, " MS)"),
         data_id = paste0(dimension, year),
         dimension = factor(dimension,
                            levels = c("Left-right", "Pro/anti-EU")))

# Reference bands so the score scale is readable without staring at
# the legend: left-right < 5 = left-of-centre majority; pro/anti-EU
# > 5 = net pro-EU majority.
p_council <- ggplot(ec_long, aes(x = year, y = score, colour = dimension)) +
  geom_hline(yintercept = 5, colour = "grey75", linewidth = 0.3) +
  geom_line(linewidth = 0.5) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = data_id),
                         size = 1.6) +
  scale_colour_manual(values = c(`Left-right`  = mo_palette$categorical[6],
                                 `Pro/anti-EU` = mo_palette$categorical[5])) +
  scale_x_continuous(breaks = seq(1990, year_now, 5)) +
  scale_y_continuous(limits = c(3, 9), breaks = c(3, 4, 5, 6, 7, 8, 9)) +
  facet_wrap(~ dimension, ncol = 1, scales = "fixed") +
  labs(x = NULL, y = "Council mean (0-10)",
       title = "Ideological composition of the Council over time",
       subtitle = paste0("Equal-weighted mean across MS of each cabinet's seat-weighted ",
                         "left-right and pro/anti-EU score"),
       colour = NULL,
       caption = paste0("Source: ParlGov (cabinets + party ideology). ",
                        "EU-stance excludes LU, CY, MT (no ParlGov eu_anti_pro). ",
                        "Grey line marks the 5/10 midpoint.")) +
  theme_mo() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", hjust = 0))

girafe(ggobj = p_council, width_svg = 9, height_svg = 5)
```

Two long-run movements show up clearly. On left-right the Council
drifted leftward in the late 1990s (the era of Blair, Schröder,
Jospin), drifted back to centre-right by the mid-2000s and has been
trending centre-right ever since, particularly post-2022. On the
EU-stance dimension the Council peaked at ≈ 8/10 in the late 1990s
and the early-2000s "Big Bang" enlargement years; it has slowly drifted
to ≈ 7.4/10 today as eurosceptic governments entered cabinet in Italy,
Slovakia and elsewhere.

# By-country breakdown

The same data, but unaggregated. Each cell below is one Member State
in one year, coloured by the seat-weighted EU stance of that year's
cabinet. The countries are ordered by their long-run average so the
most eurosceptic cluster at the top.

```{r heatmap, echo=FALSE, fig.height=7.5, fig.width=9}
country_order <- eg |>
  group_by(country) |>
  summarise(mean = mean(mean_eu_anti_pro, na.rm = TRUE), .groups = "drop") |>
  arrange(mean) |>
  pull(country)

eg_plot <- eg |>
  mutate(country = factor(country, levels = country_order),
         tooltip = paste0(country, ", ", year, ": ", cabinet_name,
                          " (EU score ",
                          ifelse(is.nan(mean_eu_anti_pro), "n/a",
                                 as.character(round(mean_eu_anti_pro, 1))),
                          "/10, left-right ",
                          round(mean_left_right, 1), "/10)"),
         data_id = paste0(iso2, year))

p_heat <- ggplot(eg_plot,
                 aes(x = year, y = country, fill = mean_eu_anti_pro,
                     tooltip = tooltip, data_id = data_id)) +
  geom_tile_interactive(colour = "white", linewidth = 0.15) +
  scale_fill_gradient2(
    low    = "#D55E00", mid = "#F0E442", high = "#0072B2",
    midpoint = 5, na.value = "grey90",
    breaks = c(0, 2.5, 5, 7.5, 10),
    labels = c("Anti-EU (0)", "2.5", "Neutral (5)", "7.5", "Pro-EU (10)"),
    limits = c(0, 10)
  ) +
  scale_x_continuous(breaks = seq(1990, year_now, 5), expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = NULL, y = NULL,
       title = "Cabinet EU stance, Member State by year",
       subtitle = "Each cell: one MS in one year, coloured by ParlGov eu_anti_pro of cabinet parties",
       fill = NULL,
       caption = paste0("Grey cells = pre-accession, post-Brexit or no ParlGov ",
                        "eu_anti_pro (Luxembourg, Cyprus, Malta).")) +
  theme_mo() +
  theme(legend.position    = "bottom",
        legend.key.width   = unit(2, "cm"),
        legend.key.height  = unit(0.4, "cm"),
        panel.grid.major   = element_blank(),
        axis.text.y        = element_text(size = 8))

girafe(ggobj = p_heat, width_svg = 9, height_svg = 7.5)
```

ParlGov scores are static per party, so the heatmap captures the
party-turnover dimension of euroscepticism but understates within-party
shifts: Fidesz, for instance, keeps its pre-Orbán "moderately pro-EU"
score throughout, even though Hungary's actual EU stance from 2010
onward is sharply different. With that caveat, the most strikingly
eurosceptic governments since 1990 include Italy's Conte-I in 2018-19,
the post-2022 Meloni cabinet and several Bulgarian and Czech
coalitions.

# Salience of "Europe" in election manifestos

The Manifesto Project codes every (quasi-)sentence of every party's
election manifesto into one of ~60 policy categories. Two are
explicitly about Europe: **per108** ("European Community/Union:
Positive") and **per110** ("European Community/Union: Negative"). The
chart below plots the EU-wide unweighted mean share of manifesto
sentences in each category by year of national elections.

```{r manifesto-trend, echo=FALSE, fig.height=4.5, fig.width=9}
em_long <- em_eu |>
  filter(year >= 1980) |>
  select(year, Positive = eu_positive_mean, Negative = eu_negative_mean) |>
  tidyr::pivot_longer(c(Positive, Negative),
                      names_to = "category", values_to = "mean_share") |>
  mutate(tooltip = paste0(year, " — ", category, ": ",
                          round(mean_share, 2), "% of manifesto sentences"),
         data_id = paste0(category, year))

p_man <- ggplot(em_long,
                aes(x = year, y = mean_share, colour = category,
                    fill = category)) +
  geom_area_interactive(aes(tooltip = tooltip, data_id = data_id),
                        alpha = 0.55, position = "identity") +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = c(Positive = mo_palette$categorical[5],
                                 Negative = mo_palette$categorical[6])) +
  scale_fill_manual(values = c(Positive = mo_palette$categorical[5],
                               Negative = mo_palette$categorical[6])) +
  scale_x_continuous(breaks = seq(1980, year_now, 5), expand = c(0, 0)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Manifesto share",
       title = "How much do EU parties talk about Europe?",
       subtitle = paste0("Annual unweighted mean of pro-EU (per108) and anti-EU (per110) ",
                         "manifesto categories across EU member-state elections"),
       colour = NULL, fill = NULL,
       caption = "Source: Manifesto Project (MPDS 2024a, static drop-in - no API pull).") +
  theme_mo()

girafe(ggobj = p_man, width_svg = 9, height_svg = 4.5)
```

Pro-EU framing rose steadily through the Maastricht / single-market
era, peaked in the late 1990s and has since drifted down as openly
euro-critical campaigning has spread across the continent. What's
measured is _attention_, not endorsement; the rise in negative coverage
after 2010 reflects both genuinely eurosceptic parties and pro-EU
parties responding to them.

# Cite

Ovádek, M. *euverse: Politics of EU Member State Governments*. Accessed
`r format(today_d, "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-politics.html>.
