---
title: "Politics of EU Member State Governments"
subtitle: "Ideological and pro/anti-EU composition of national cabinets, plus the EU as a campaign issue"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggiraph)
  library(here)
  library(stringr)
  library(forcats)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Pre-built parquets from R/build_eu_politics.R - regenerated nightly
# from ParlGov view_cabinet + view_party joined to data-manual/eu-member-states
# and from the Manifesto Project MPDS dump.
eg <- arrow::read_parquet(
  here::here("data-final", "eu_government_composition.parquet")
)
em_eu <- arrow::read_parquet(
  here::here("data-final", "eu_manifesto_eu_stance_eu.parquet")
)
em_co <- arrow::read_parquet(
  here::here("data-final", "eu_manifesto_eu_stance_country.parquet")
)

# Freshness signals come from the build sidecars (these are derived
# data, not API snapshots, so use manual_meta).
eg_meta <- manual_meta(
  here::here("data-final", "eu_government_composition.parquet"),
  source_label = "ParlGov view_cabinet + view_party (built nightly)"
)
eg_meta$rows <- nrow(eg)
em_meta <- manual_meta(
  here::here("data-final", "eu_manifesto_eu_stance_eu.parquet"),
  source_label = "Manifesto Project MPDS 2024a"
)
em_meta$rows <- nrow(em_co)

today_d  <- Sys.Date()
year_now <- as.integer(format(today_d, "%Y"))
```

The euverse's other trackers describe what EU governments _do_ - vetoes
they cast, laws they pass, cases they litigate. This page is about
_who they are_: the partisan and ideological composition of each
Member State's national government over time, and how openly the
parties making up those governments support (or oppose) European
integration.

Two complementary data sources sit underneath:

- **[ParlGov](https://www.parlgov.org/)** for the factual record of
  who has been in cabinet, when, and with what share of seats. ParlGov
  also assigns each party fixed left-right and pro/anti-EU scores
  (`eu_anti_pro`, 0 = anti-EU, 10 = pro-EU) drawn from expert surveys.
- **The [Manifesto Project (MPDS)](https://manifesto-project.wzb.eu/)**
  for what parties actually wrote about Europe in their election
  manifestos, decade by decade.

```{r freshness, results='asis', echo=FALSE}
freshness_badge_combined(list(eg_meta, em_meta))
```

# Pro/anti-EU stance of governments

The heatmap below shows the seat-weighted average pro/anti-EU score of
each Member State's national cabinet for every year since 1990. Each
cell is one country-year: warm colours mean a strongly pro-EU
government, cool colours an openly eurosceptic one. ParlGov scores
parties on the 0-10 scale and we weight by parliamentary seats, so a
cabinet whose biggest partner is mildly eurosceptic but whose junior
partner is strongly pro-EU averages to "neutral".

```{r heatmap, echo=FALSE, fig.height=7.5, fig.width=9}
# Order countries by their long-run mean eu_anti_pro score so the most
# eurosceptic cabinets cluster near the top.
country_order <- eg |>
  group_by(country) |>
  summarise(mean = mean(mean_eu_anti_pro, na.rm = TRUE), .groups = "drop") |>
  arrange(mean) |>
  pull(country)

eg_plot <- eg |>
  mutate(country = factor(country, levels = country_order),
         tooltip = paste0(country, ", ", year, ": ", cabinet_name,
                          " (EU score ", round(mean_eu_anti_pro, 1),
                          "/10, left-right ", round(mean_left_right, 1), "/10)"),
         data_id = paste0(iso2, year))

p_heat <- ggplot(eg_plot,
                 aes(x = year, y = country, fill = mean_eu_anti_pro,
                     tooltip = tooltip, data_id = data_id)) +
  geom_tile_interactive(colour = "white", linewidth = 0.15) +
  scale_fill_gradient2(
    low    = "#D55E00", mid = "#F0E442", high = "#0072B2",
    midpoint = 5, na.value = "grey90",
    breaks = c(0, 2.5, 5, 7.5, 10),
    labels = c("Anti-EU (0)", "2.5", "Neutral (5)", "7.5", "Pro-EU (10)"),
    limits = c(0, 10)
  ) +
  scale_x_continuous(breaks = seq(1990, year_now, 5),
                     expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0)) +
  labs(x = NULL, y = NULL,
       title = "Seat-weighted EU stance of national cabinets",
       subtitle = "Each cell: one Member State in one year, coloured by ParlGov eu_anti_pro of cabinet parties",
       fill = NULL,
       caption = "Countries ordered by long-run average. Blank cells = pre-accession (or post-Brexit, for the UK).") +
  theme_mo() +
  theme(legend.position    = "bottom",
        legend.key.width   = unit(2, "cm"),
        legend.key.height  = unit(0.4, "cm"),
        panel.grid.major   = element_blank(),
        axis.text.y        = element_text(size = 8))

girafe(ggobj = p_heat, width_svg = 9, height_svg = 7.5)
```

ParlGov scores are static per party, so the heatmap captures party
turnover in cabinet but not within-party shifts: Fidesz, for instance,
keeps its pre-Orbán "moderately pro-EU" score throughout, understating
the country's actual euroscepticism since 2010. With that caveat in
mind, the most strikingly eurosceptic governments since 1990 include
Italy's Conte-I (Lega + M5S) in 2018-19, the post-2022 Meloni cabinet,
Slovakia under Fico, and several Bulgarian and Czech coalitions.

# Eurosceptic parties in cabinet

A complementary, less aggregated view: count how many Member States in
each year had _at least one_ cabinet party scoring below 4 on ParlGov's
`eu_anti_pro` scale - the standard threshold for openly eurosceptic
party families.

```{r euroscept-count, echo=FALSE, fig.height=4.5, fig.width=9}
euroscept_by_year <- eg |>
  group_by(year) |>
  summarise(
    n_total          = dplyr::n(),
    n_with_eurosc    = sum(has_eurosceptic, na.rm = TRUE),
    share_with_eurosc = n_with_eurosc / n_total,
    .groups          = "drop"
  ) |>
  mutate(tooltip = paste0(year, ": ", n_with_eurosc, " of ", n_total,
                          " MS governments include a eurosceptic party"),
         data_id = as.character(year))

p_es <- ggplot(euroscept_by_year,
               aes(x = year, y = n_with_eurosc)) +
  geom_col_interactive(aes(tooltip = tooltip, data_id = data_id),
                       fill = mo_palette$categorical[6],
                       colour = "white", linewidth = 0.2) +
  scale_x_continuous(breaks = seq(1990, year_now, 5), expand = c(0, 0)) +
  scale_y_continuous(breaks = int_breaks(), expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Member States",
       title = "EU governments including at least one eurosceptic party",
       subtitle = paste0("Counted across all current and historical EU MS in each year ",
                         "(threshold: a cabinet party with ParlGov eu_anti_pro <= 4)"),
       caption = "Source: ParlGov + EU MS master.") +
  theme_mo()

girafe(ggobj = p_es, width_svg = 9, height_svg = 4.5)
```

# Salience of "Europe" in election manifestos

The Manifesto Project codes every sentence (quasi-sentence) of every
party's election manifesto into one of ~60 policy categories. Two of
those categories are explicitly about Europe: **per108**
("European Community/Union: Positive") and **per110**
("European Community/Union: Negative"). The chart below plots the
EU-wide unweighted mean share of manifesto sentences in each category
per year of national elections, plus the running net (positive minus
negative).

```{r manifesto-trend, echo=FALSE, fig.height=4.5, fig.width=9}
em_long <- em_eu |>
  filter(year >= 1980) |>
  select(year, Positive = eu_positive_mean, Negative = eu_negative_mean) |>
  tidyr::pivot_longer(c(Positive, Negative),
                      names_to = "category", values_to = "mean_share") |>
  mutate(tooltip = paste0(year, " - ", category, ": ",
                          round(mean_share, 2), "% of manifesto sentences"),
         data_id = paste0(category, year))

p_man <- ggplot(em_long,
                aes(x = year, y = mean_share, colour = category,
                    fill = category)) +
  geom_area_interactive(aes(tooltip = tooltip, data_id = data_id),
                        alpha = 0.55, position = "identity") +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = c(Positive = mo_palette$categorical[5],
                                 Negative = mo_palette$categorical[6])) +
  scale_fill_manual(values = c(Positive = mo_palette$categorical[5],
                               Negative = mo_palette$categorical[6])) +
  scale_x_continuous(breaks = seq(1980, year_now, 5), expand = c(0, 0)) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Manifesto share",
       title = "How much do EU parties talk about Europe?",
       subtitle = paste0("Annual unweighted mean of pro-EU (per108) and anti-EU (per110) ",
                         "manifesto categories across EU member-state elections"),
       colour = NULL, fill = NULL,
       caption = "Source: Manifesto Project (MPDS 2024a).") +
  theme_mo()

girafe(ggobj = p_man, width_svg = 9, height_svg = 4.5)
```

The series tells a familiar story: pro-EU framing rose steadily through
the Maastricht / single-market era, peaked in the late 1990s, and has
since drifted down as openly euro-critical campaigning has spread
across the continent. Note that what's measured is _attention_, not
endorsement; the rise in negative coverage after 2010 reflects both
genuinely eurosceptic parties and pro-EU parties responding to them.

# Country-level EU stance over time

The same Manifesto-project measure, but split out by Member State. A
country whose line drops below zero is one whose election manifestos
that year, on average, said more negative things about the EU than
positive ones.

```{r manifesto-country, echo=FALSE, fig.height=8, fig.width=9}
# Restrict to (a) years where multiple parties wrote manifestos, so the
# means aren't dominated by a single small party, and (b) a recent
# window where the EU was already a contested issue.
em_co_plot <- em_co |>
  filter(n_parties >= 3, year >= 1995)

# Eurosceptic countries that the user-facing reader is most likely to
# care about - shown first (top of facet grid).
notable <- c("United Kingdom", "Hungary", "Poland", "Italy", "France",
             "Germany", "Greece", "Netherlands", "Denmark", "Sweden",
             "Austria", "Czechia", "Slovakia", "Spain", "Finland", "Ireland")
em_co_plot <- em_co_plot |>
  filter(country %in% notable) |>
  mutate(country = factor(country, levels = notable),
         tooltip = paste0(country, " ", year, ": net ",
                          round(net_pro_eu, 2),
                          "% (", round(eu_positive, 2), " positive, ",
                          round(eu_negative, 2), " negative)"),
         data_id = paste0(iso2, year))

p_co <- ggplot(em_co_plot,
               aes(x = year, y = net_pro_eu)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.3) +
  geom_line(colour = mo_palette$categorical[2], linewidth = 0.4) +
  geom_point_interactive(aes(tooltip = tooltip, data_id = data_id,
                             colour = net_pro_eu),
                         size = 1.6) +
  scale_colour_gradient2(low = mo_palette$categorical[6],
                         mid = "grey70",
                         high = mo_palette$categorical[5],
                         midpoint = 0, guide = "none") +
  scale_x_continuous(breaks = c(1995, 2005, 2015, 2025)) +
  facet_wrap(~ country, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = "Net pro-EU manifesto share",
       title = "EU as an election issue, by Member State",
       subtitle = "Vote-weighted mean per108 - per110 per election year; negative = net hostile",
       caption = "Source: Manifesto Project. Years with <3 parties coded omitted.") +
  theme_mo() +
  theme(strip.text = element_text(face = "bold", hjust = 0))

girafe(ggobj = p_co, width_svg = 9, height_svg = 8)
```

# Cite

Ovádek, M. *Politics of EU Member State Governments*. Accessed
`r format(today_d, "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-politics.html>.
