---
title: "Tracking EU Member States' Vetoes"
subtitle: "Who and when blocks joint EU action"
toc: true
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggiraph)
  library(here)
})
source(here::here("R", "theme.R"))
source(here::here("R", "freshness.R"))
source(here::here("R", "table.R"))
register_ggiraph_defaults()

fmt_n <- function(x) format(x, big.mark = ",")

# Load the manual data.
veto_csv <- here::here("data-manual", "ms-vetoes.csv")
veto_data <- read.csv(veto_csv, sep = ";", fileEncoding = "UTF-8-BOM") |>
  # Drop columns that are all-empty (legacy CSV trailing-comma artefact).
  select(where(~ !(all(is.na(.)) | all(. == "")))) |>
  filter(as.Date(date_veto) > as.Date("2011-01-01")) |>
  arrange(date_veto)

# Add IDs (matches old Rmd structure).
veto_data <- veto_data |>
  mutate(id_veto = row_number(), .before = 1) |>
  group_by(date_veto, issue, issue_country, source) |>
  mutate(id_issue = cur_group_id(), .after = 1) |>
  ungroup()

# Freshness meta: use the date of the LATEST reported veto rather than
# the CSV's mtime. CI checkout resets file mtimes to the run time, so
# mtime would show "fresh" every nightly render even if nobody added a
# new veto in months — misleading. The most-recent veto date is the
# honest signal for "is this dataset current?".
latest_veto_date <- max(as.Date(veto_data$date_veto), na.rm = TRUE)
veto_meta <- list(
  source     = "data-manual/ms-vetoes.csv (hand-curated)",
  fetched_at = format(as.POSIXct(latest_veto_date), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
  rows       = nrow(veto_data)
)

# Inline-prose stats.
start_date <- format(min(as.Date(veto_data$date_veto)), "%d %B %Y")
n_vetoes   <- nrow(veto_data)
n_states   <- length(unique(veto_data$ms_veto))
n_issues   <- max(veto_data$id_issue)
```

This page tracks all publicly reported instances of European Union Member
States vetoing joint action. A veto is understood as an instance of a
Member State blocking, temporarily or permanently, common EU action by
opposing a measure under a procedure requiring unanimous agreement of all
Member States.^[I count every reported instance of a veto being used, even
if the same or similar issue already attracted a veto in the past.]

Since `r start_date`, there have been `r mono(fmt_n(n_vetoes))` vetoes by
`r mono(n_states)` Member States across `r mono(n_issues)` issues. This
page is updated whenever a new veto is reported.

```{r freshness, results='asis', echo=FALSE}
freshness_badge(veto_meta)
```

```{r vetoes-table, echo=FALSE}
veto_data |>
  arrange(desc(as.Date(date_veto))) |>
  transmute(
    Date    = as.Date(date_veto),
    Vetoer  = ms_veto,
    Issue   = issue,
    Country = ifelse(issue_country == "", "—", issue_country),
    Source  = sprintf('<a href="%s" target="_blank">link</a>', source)
  ) |>
  dt_mo(escape = FALSE)
```

# Vetoes by Member State

Hungary has been reported to have vetoed EU action more than any other
Member State in recent history. In contrast, quite a few Member States
have not had a single veto attributed to them. However, by relying on
reported vetoes only, this statistic likely underestimates the true
prevalence of vetoes. Another source of underestimation are long-standing
vetoes: certain issues might not even be put on the Council agenda — and
are therefore less likely to be reported — when one or several countries
had vetoed the policy in question in the past.^[An example here is the
opening of accession negotiations between the EU and (North) Macedonia,
which was blocked by Greece for over a decade until a dispute over the
country's name was resolved.]

```{r plot-by-ms, echo=FALSE, fig.height=4.5, fig.width=9}
df_a <- veto_data |>
  count(ms_veto) |>
  mutate(fill_hex = unname(mo_palette$ms_flags[ms_veto]))

p_a <- ggplot(df_a, aes(x = reorder(ms_veto, n), y = n,
                        fill = fill_hex,
                        tooltip = paste0(ms_veto, ": ", n, " vetoes"),
                        data_id = ms_veto)) +
  geom_col_interactive(show.legend = FALSE) +
  scale_fill_identity() +
  scale_y_continuous(n.breaks = 8, expand = c(0, 0.1)) +
  labs(x = NULL, y = "Total number of vetoes",
       title = "Number of vetoes by Member State") +
  theme_mo() +
  theme(panel.grid = element_blank(),
        panel.grid.major.y = element_line(colour = "grey95"),
        axis.text.x = element_text(face = "bold"))

girafe(ggobj = p_a, width_svg = 9, height_svg = 4.5)
```

# Vetoes over time

Reported vetoes have become more common over time.^[The date of the veto
is an approximation of when the negotiation was taking place rather than
when the veto was reported.] Some would argue this is a signal of
increasing polarization in the Council of the EU. But changes to the EU's
institutional system also play a role. Notably, the creation of the office
of the High Representative for Foreign and Security Affairs by the Lisbon
Treaty has generated more opportunities for Member States to veto joint
EU statements on foreign policy issues.

This is also a reminder that not all vetoes are the same — blocking a
legislative measure such as the harmonization of minimum corporate tax is
arguably much more consequential than the wording of a non-binding
declaration.

```{r plot-over-time, echo=FALSE, fig.height=6, fig.width=9}
df_b <- veto_data |>
  mutate(
    hex_color = unname(mo_palette$ms_flags[ms_veto]),
    tooltip   = paste0("By: ", ms_veto, "\n",
                       "On: ", format(as.Date(date_veto), "%d %B %Y"), "\n",
                       "Issue: ", issue)
  )

p_b <- ggplot(df_b, aes(x = as.Date(date_veto), y = id_veto)) +
  geom_vline(xintercept = as.Date(c("2014-12-01", "2019-12-01", "2024-12-01")),
             lty = 2, colour = "grey60") +
  annotate("text", label = "Juncker", angle = 90, size = 3, colour = "grey50",
           x = as.Date("2014-10-01"), y = -Inf, vjust = 0, hjust = 0) +
  annotate("text", label = "von der Leyen I", angle = 90, size = 3, colour = "grey50",
           x = as.Date("2019-10-01"), y = -Inf, vjust = 0, hjust = 0) +
  annotate("text", label = "von der Leyen II", angle = 90, size = 3, colour = "grey50",
           x = as.Date("2024-10-01"), y = -Inf, vjust = 0, hjust = 0) +
  geom_point_interactive(aes(colour = hex_color,
                             tooltip = tooltip,
                             data_id = id_veto),
                         alpha = 0.9) +
  scale_color_identity() +
  scale_x_date(date_breaks = "year", date_labels = "%Y") +
  scale_y_continuous(breaks = seq(0, n_vetoes, by = 5)) +
  theme_mo() +
  theme(axis.text.y = element_text(size = 6)) +
  labs(x = NULL, y = "Veto No",
       title = "Vetoes over time",
       caption = "Hover for more information about the veto")

girafe(ggobj = p_b, width_svg = 9, height_svg = 6)
```

When trying to interpret these data, one should be wary of the
possibility that differences over time and between Member States are to
some extent driven by how vetoes are communicated and reported.

Moreover, whether countries choose to exercise a veto power can also be a
function of other Member States' intentions to do the same. If Member
State A knows for certain that Member State B will veto a measure, it can
reap the 'benefit' of the failed action without contributing to it
(free-riding). The number of reported vetoes therefore understates the
true level of disagreement among Member States.

# Veto issues and third countries

We can also categorize the vetoes by what substantive issue they
concerned and — because vetoes are frequently about foreign affairs —
the third country in question. Multiple vetoes have related to progress
in negotiating the accession of a candidate country outside the EU. Not
all vetoes are directly linked to a third country; this is especially
true when it comes to internal policies such as taxation.

```{r plot-issues, echo=FALSE, fig.height=5, fig.width=9}
df_c <- veto_data |>
  mutate(issue_country = ifelse(issue_country == "", "No third country", issue_country)) |>
  count(issue, issue_country) |>
  mutate(sum = sum(n), .by = issue) |>
  mutate(
    tooltip = ifelse(issue_country == "No third country",
                     paste0(n, " vetoes concerning ", issue),
                     paste0(n, " vetoes concerning ", issue, " and ", issue_country)),
    id = seq_len(dplyr::n())
  )

p_c <- ggplot(df_c, aes(x = n, y = reorder(issue, sum))) +
  geom_col_interactive(aes(fill = issue_country,
                           tooltip = tooltip,
                           data_id = id),
                       show.legend = FALSE) +
  ggiraph::scale_fill_brewer_interactive(palette = "Paired", type = "qual") +
  # Add a small left-side data padding so bars don't start flush against
  # the y-axis labels; visually separates the label column from the bars.
  scale_x_continuous(expand = expansion(mult = c(0.04, 0.04))) +
  theme_mo() +
  # Smaller, right-aligned y labels with explicit tick padding. The
  # element_text margin alone wasn't shifting positions in ggiraph's
  # SVG; combining with a smaller text size makes the label column
  # visibly less cramped.
  theme(axis.text.y      = element_text(hjust = 1, size = 9),
        axis.ticks.length = unit(4, "pt")) +
  labs(x = NULL, y = NULL,
       title = "Veto issues and third countries",
       subtitle = "Number of times a veto related to an issue",
       caption = "Hover for information about the third country")

girafe(ggobj = p_c, width_svg = 9, height_svg = 5)
```

# Cite

Ovádek, M. *Tracking EU Member States' Vetoes*. Accessed
`r format(Sys.Date(), "%d %B %Y")`. Available at
<https://michalovadek.github.io/euverse/trackers/eu-vetoes.html>.
